---
title: Differences in care for patients experiencing left vs. right hemisphere ICH
author: Nikhil Patel, MD
format: html
toc: true
---


```{r setup, output=FALSE, message=FALSE, warning=FALSE}
#| code-fold: false

library(targets)
library(readr)
library(tidyr)
library(janitor)
library(dplyr)
library(forcats)
library(stringr)
library(skimr)
library(gt)
library(gtsummary)
library(marginaleffects)
library(modelsummary)

tar_load(ich)
tar_load(non_surgery_trials)
tar_load(surgery_trials)
tar_load(table_one)

```

## Exploratory Data Analysis

::: {.callout-warning title="Issues to Address"}
`htn` is missing mostly all information. It is also listed as a character variable.

All other co-morbidities are missing information; if this is the case, should default to "No".

`ivh` is also missing some values, which should also default to "No"?

Any "Yes/No" variable should not have any missing information, should just default to "No"

Verify accuracy/validity of `time_symptoms_to_ed`, `sbp_baseline`, 

Fix/drop unused levels in `race`, and fix "American Indian"

Fix "unknown" category in `ethnicity` to "Not Hispanic or Latino"

Fix/drop unused levels in `ich_location` and `study`
:::

```{r}

glimpse(ich)
skim(ich)

```

## Table 1

::: {.callout-warning title="Issues to Address"}

Change order of the displayed to frequency

:::

```{r}

table_one

```

