---
title: Patients with Left Hemispheric Intracerebral Hemorrhage are Less Likely to Receive Neurosurgical Intervention
author: "Nikhil Patel, MD"
subtitle: Supplementary Appendix
authors:
  - name: Nikhil Patel
    orcid: 0000-0002-3193-8590
    affiliation: Department of Internal Medicine, Wake Forest University School of Medicine
    roles: data analysis, writing
    corresponding: true
  - name: Matt Jaffa
    affiliation: Department of Neurology, University of Connecticut School of Medicine
  - name: Rahul Karamchandani
    affiliation: Department of Neurology, Wake Forest University School of Medicine
  - name: Andrew Asimos
    affiliation: Department of Emergency Medicine, Wake Forest University School of Medicine
  - name: Jonathan Parish
    affiliation: Carolina Neurosurgery & Spine Associates
  - name: Nicholas Allen Morris
    affiliation: Department of Neurology, University of Maryland School of Medicine
date: last-modified
abstract: |
  
table-of-contents: true
toc-title: Contents
---

```{r setup}
#| echo: false
#| output: false

library(targets)
library(tidyverse)
library(withr)
library(here)
library(modelsummary)
library(marginaleffects)
library(brms)
library(tidybayes)
library(bayesplot)
library(glue)
library(ggdist)
library(patchwork)
library(tinytable)
library(flextable)


# plot stuff
theme_ich <- function(base_family = "Figtree",
                      base_size = 14) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(lineheight = 2),
      plot.title = element_text(size = rel(1.3), margin = margin(12, 0, 8, 0)),
      plot.subtitle = element_text(size = rel(1.1), margin = margin(4, 0, 0, 0)),
      axis.text.y = element_text(size = rel(0.8)),
      axis.title.y = element_text(size = 12, margin = margin(0, 4, 0, 0)),
      axis.text.x = element_text(size = rel(2)),
      axis.title.x = element_blank(),
      legend.position = "bottom",
      legend.justification = 1,
      panel.grid = element_line(colour = "#F3F4F5"),
      plot.caption = element_text(size = rel(0.8), margin = margin(8, 0, 0, 0)),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25, "cm")
    )
}

# visualization function
results_visual <- function(model, title) {
  model <- model
  subtitle <- title

  draws <- model |>
    spread_draws(b_ich_lateralityRight) |>
    mutate(ich_right_or = exp(b_ich_lateralityRight)) |>
    summarize()

  pred <- predictions(model, by = "ich_laterality") |>
    tibble()

  theme_ich <- function(base_family = "Inter",
                        base_size = 14) {
    theme_minimal(base_size = base_size) +
      theme(
        text = element_text(lineheight = 2),
        plot.title = element_text(size = rel(1.3), margin = margin(12, 0, 8, 0)),
        plot.subtitle = element_text(size = rel(1.1), margin = margin(4, 0, 0, 0)),
        axis.text.y = element_text(size = rel(0.8)),
        axis.title.y = element_text(size = 12, margin = margin(0, 4, 0, 0)),
        axis.text.x = element_text(size = rel(2)),
        axis.title.x = element_blank(),
        legend.position = "bottom",
        legend.justification = 1,
        panel.grid = element_line(colour = "#F3F4F5"),
        plot.caption = element_text(size = rel(0.8), margin = margin(8, 0, 0, 0)),
        plot.margin = margin(0.25, 0.25, 0.25, 0.25, "cm")
      )
  }

  p1 <- predictions(model, by = "ich_laterality") |>
    posterior_draws() |>
    ggplot(aes(x = draw, fill = ich_laterality)) +
    geom_density(alpha = .7) +
    scale_fill_manual(values = c("#422f05", "#E69F00")) +
    scale_x_continuous(labels = scales::percent) +
    labs(
      title = "Predicted Probability",
      fill = "ICH Laterality",
      y = NULL
    ) +
    theme_ich()

  p2 <- avg_comparisons(model, variables = "ich_laterality") |>
    posterior_draws() |>
    ggplot(aes(x = draw)) +
    geom_density(color = "#422f05", fill = "#E69F00") +
    scale_x_continuous(labels = scales::percent) +
    labs(
      title = "Average Difference between Left vs. Right Hemisphere",
      y = NULL
    ) +
    geom_vline(
      xintercept = 0, linetype = "dashed",
      color = "#999999", linewidth = 1
    ) +
    theme_ich()

  p3 <- model |>
    spread_draws(b_ich_lateralityRight) |>
    mutate(ich_right_or = exp(b_ich_lateralityRight)) |>
    ggplot(aes(ich_right_or)) +
    geom_density(color = "#422f05", fill = "#E69F00") +
    # stat_halfeye(aes(fill_ramp = stat(x > 1), fill = "#E69F00", color = "#422f05")) +
    scale_x_continuous(breaks = seq(0, 3, 0.25)) +
    labs(
      title = "Odds Ratio - Right Hemisphere",
      y = NULL,
      fill = NULL
    ) +
    geom_vline(
      xintercept = 1, linetype = "dashed",
      color = "#999999", linewidth = 1
    ) +
    theme_ich() +
    theme(legend.position = "none")


  total <- p1 / p2 / p3

  final <- total +
    plot_annotation(
      title = glue(subtitle),
      theme = theme(plot.title = element_text(size = rel(2), margin = margin(12, 0, 8, 0)))
    )

  return(final)
}


# load data and models
with_dir(here(), tar_load(ich_aggressive))

```

# Details of Statistical Analysis

## Covariate Selection

```{r dags}
#| echo: false


# dag_aggressive

```

## Prior Distribution Definitions

```{r prior definitions}
#| echo: false
#| output: false

neutral_mean <- 0
neutral_sd <- 0.35
neutral_lower_OR <- exp(neutral_mean - (1.96 * neutral_sd))
neutral_upper_OR <- exp(neutral_mean + (1.96 * neutral_sd))

left_mean <- -0.22
left_sd <- 0.175
left_lower_OR <- exp(left_mean - (1.96 * left_sd))
left_upper_OR <- exp(left_mean + (1.96 * left_sd))

right_mean <- 0.18
right_sd <- 0.175
right_lower_OR <- exp(right_mean - (1.96 * right_sd))
right_upper_OR <- exp(right_mean + (1.96 * right_sd))

```

In defining prior distributions, we generally followed suggestions from previous literature. [@zampieriUsingBayesianMethods2021; @kruschkeBayesianAnalysisReporting2021; @gelmanBayesianWorkflow2020] Specifically, we defined 4 sets of priors for each outcome:

**Neutral prior**: mean of **`r neutral_mean`** and standard deviation of **`r neutral_sd`**, centering around an OR of **`r exp(neutral_mean)`** and 95% of the the probability mass for the OR between **`r round(neutral_lower_OR, digits = 2)`** to **`r round(neutral_upper_OR, digits = 2)`**.
    
**Left hemisphere prior**: mean of **`r left_mean`** and standard deviation of **`r left_sd`**, centering around an OR of **`r round(exp(left_mean), digits = 2)`** and 95% of the the probability mass for the OR between **`r round(left_lower_OR, digits = 2)`** to **`r round(left_upper_OR, digits = 2)`**.

**Right hemisphere prior**: mean of **`r right_mean`** and standard deviation of **`r right_sd`**, centering around an OR of **`r round(exp(right_mean), digits =2)`** and 95% of the the probability mass for the OR between **`r round(right_lower_OR, digits = 2)`** to **`r round(right_upper_OR, digits =2)`**.

**Flat prior**: most mimics a frequentist analysis
    
```{r prior graphical summary}
#| echo: false
#| fig-height: 6

# Return dnorm(x) for 0 < x < 2, and NA for all other x
dnorm_limit <- function(x) {
  y <- dnorm(x)
  y[x < 0] <- NA
  return(y)
}

ggplot(data = tibble(x = -2:2), aes(x = x)) +
  geom_function(fun = dnorm, args = list(mean = -20, sd = 0.5), linewidth = 1) +
  # stat_function(fun = dnorm_limit, geom = "area", color = "orange", alpha = 0.2) +
  geom_vline(xintercept = -0.22, linetype = "dashed", color = "orange", linewidth = 1) +
  scale_x_continuous(breaks = seq(-2, 2, 0.5)) +
  theme_ich()

```

### Neurosurgical Intervention

```{r informative prior predictive check - neurosurgical intervention}
#| echo: false
#| fig-height: 12
#|

with_dir(here(), tar_load(m_prior_neutral_neurosurgery))
with_dir(here(), tar_load(m_prior_left_neurosurgery))


avg_predictions(m_prior_left_neurosurgery, variables = "ich_laterality")

results_visual(m_prior_left_neurosurgery, "Prior Predictive Check")

```

## Posterior Simulation Diagnostics

### MCMC Chain Convergence and Resolution

### Posterior predictive checks

### 

## Sensitivity Analyses

