---
title: "Association of Hemispheric Laterality with Neurosurgical Intervention in Intracerebral Hemorrhage: A Post-Hoc Bayesian Analysis of ATACH-2 and ERICH"
authors:
  - name: Nikhil M. Patel
    degrees: MD, MBA
    orcid: 0000-0002-3193-8590
    affiliations:
      - ref: wfsom-int-med
      - ref: wfsom-neuro
    corresponding: true
    email: nikhil.m.patel@wfusm.edu
  - name: Rahul R. Karamchandani
    degrees: MD
    affiliations:
      - ref: wfsom-neuro
  - name: Matthew N. Jaffa
    degrees: DO
    affiliations:
      - ref: hartford-neuro
  - name: Liang Wang
    degrees: PhD
    affiliations:
      - ref: advocate-analytics
  - name: Hongmei Yang
    degrees: PhD
    affiliations:
      - ref: advocate-analytics
  - name: Rhonda Cadena
    degrees: MD
    affiliations:
      - ref: wfsom-int-med
  - name: Christa Swisher
    degrees: MD
    affiliations:
      - ref: wfsom-neuro
  - name: Joe D. Bernard
    degrees: MD
    affiliations:
      - ref: cnsa
  - name: Jonathan M. Parish
    degrees: MD
    affiliations:
      - ref: cnsa
  - name: William R. Stetler
    degrees: MD
    affiliations:
      - ref: cnsa
  - name: Rick Magun
    degrees: MD
    affiliations:
      - ref: buffalo-neuro
  - name: Aaron Taylor
    degrees: MD
    affiliations:
      - ref: buffalo-neuro
  - name: Daniel Woo
    degrees: MD
    affiliations:
      - ref: buffalo-neuro
  - name: Adnan Qureshi
    degrees: MD
    affiliations:
      - ref: missouri-neuro
  - name: Andrew W. Asimos
    degrees: MD
    affiliations:
      - ref: wfsom-em
  - name: Nicholas A. Morris
    degrees: MD
    affiliations:
      - ref: umaryland-neuro

affiliations:
  - id: wfsom-int-med
    name: Department of Internal Medicine, Wake Forest School of Medicine, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: wfsom-neuro
    name: Department of Neurology, Neurosciences Institute, Wake Forest School of Medicine, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: hartford-neuro
    name: Division of Neurocritical Care, Department of Neurology, Ayer Neuroscience Institute, Hartford Hospital
    city: Hartford
    state: CT
    country: USA
  - id: advocate-analytics
    name: Clinical Quality Analytics, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: cnsa
    name: Carolina Neurosurgery & Spine Associates, Department of Neurosurgery, Neurosciences Institute, Wake Forest University School of Medicine, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: buffalo-neuro
    name: Department of Neurology, University at Buffalo Jacobs School of Medicine & Biomedical Sciences
    city: Buffalo
    state: NY
    country: USA
  - id: missouri-neuro
    name: Zeenat Qureshi Stroke Institute and Department of Neurology, University of Missouri-Columbia
    city: Columbia
    state: MO
    country: USA
  - id: wfsom-em
    name: Department of Emergency Medicine, Neurosciences Institute, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: umaryland-neuro
    name: Department of Neurology, Program in Trauma, University of Maryland School of Medicine
    city: Baltimore
    state: MD
    country: USA
format:
  docx:
    reference-doc: custom-reference.docx
bibliography: references.bib
csl: stroke.csl
table-of-contents: false
toc-title: Contents
echo: false
output: false
message: false
---

```{r setup}
library(targets)
library(tidyverse)
library(scales)
library(withr)
library(here)
library(gt)

# load data
with_dir(here(), tar_load(ich_aggressive))

## patient counts
sample_size <- count(ich_aggressive)

left_hemisphere_count <- ich_aggressive |>
  filter(ich_laterality == "Left") |>
  count()

erich_count <- ich_aggressive |>
  filter(study == "ERICH") |>
  count()

atach_count <- ich_aggressive |>
  filter(study == "ATACH-2") |>
  count()

## table 1
with_dir(here(), tar_load(table_1))

## table 2
with_dir(here(), tar_load(table_2_neutral))
tibble2 <- table_2_neutral[["_data"]]

## table 3
with_dir(here(), tar_load(table_3_neutral))
tibble3 <- table_3_neutral[["_data"]]
```

```{r neurosurgical-intervention}
# Extract the values for the neurosurgical intervention outcome
or_ci_values_surg <- tibble2 %>%
  filter(Outcome == "Neurosurgical Intervention") %>%
  pull(or_ci)

or_ci_split_surg <- str_split(or_ci_values_surg, " \\(| - ", simplify = TRUE)

odds_ratio_surgery <- or_ci_split_surg[[1]]
lower_bound_surg <- or_ci_split_surg[[2]]
upper_bound_surg <- str_remove(or_ci_split_surg[[3]], "\\)")
```

```{r secondary-outcomes}
# EVD
or_ci_values_evd <- tibble2 %>%
  filter(Outcome == "External Ventricular Drain") %>%
  pull(or_ci)

or_ci_split_evd <- str_split(or_ci_values_evd, " \\(| - ", simplify = TRUE)

odds_ratio_evd <- or_ci_split_evd[[1]]
lower_bound_evd <- or_ci_split_evd[[2]]
upper_bound_evd <- str_remove(or_ci_split_evd[[3]], "\\)")

# Mechanical ventilation (count outcome; reported as IRR)
irr_ci_values_mv <- tibble2 %>%
  filter(Outcome == "Days of Mechanical Ventilation") %>%
  pull(or_ci)

irr_ci_split_mv <- str_split(irr_ci_values_mv, " \\(| - ", simplify = TRUE)

irr_mv <- irr_ci_split_mv[[1]]
lower_bound_mv <- irr_ci_split_mv[[2]]
upper_bound_mv <- str_remove(irr_ci_split_mv[[3]], "\\)")
```

## Abstract

**Background**: Hemispheric laterality influences clinical management in stroke. Given the increasing enthusiasm for surgical management of intracerebral hemorrhage (ICH), we evaluated the association of hemispheric lateralization with the likelihood of receiving neurosurgical intervention and subsequent 90-day functional outcomes.

**Methods**: This post-hoc analysis included `r sample_size %>% pull(n)` patients with supratentorial ICH from the ATACH-2 and ERICH studies. The primary outcome was neurosurgical intervention (hematoma evacuation, decompressive craniectomy, or other craniotomy). Secondary outcomes included external ventricular drain (EVD) placement, 90-day modified Rankin Scale (mRS), and EuroQOL domains. Hierarchical Bayesian regression models were adjusted for age, admission Glasgow Coma Scale, ICH location, ICH volume, and intraventricular hemorrhage.

**Results**: Of `r sample_size %>% pull(n)` patients, right hemispheric ICH was associated with a higher likelihood of receiving neurosurgical intervention (aOR `r odds_ratio_surgery`, 95% CrI `r lower_bound_surg`–`r upper_bound_surg`) and EVD placement (aOR `r odds_ratio_evd`, 95% CrI `r lower_bound_evd`–`r upper_bound_evd`) compared to left hemispheric ICH. This association was directionally consistent across ICH locations, study cohorts, and prior specifications. Despite the lower rates of aggressive surgical management, patients with left hemispheric ICH demonstrated similar 90-day functional outcomes, exhibiting equivalent mRS scores and potentially better outcomes in specific health-related quality of life domains.

**Conclusions**: Right hemispheric ICH is associated with a higher likelihood of neurosurgical intervention compared to left hemispheric ICH, though 90-day functional outcomes remain similar between groups. While these findings highlight a potential laterality-driven disparity in treatment paradigms, residual confounding from unmeasured anatomical severity and family preferences cannot be excluded. Prospective studies are needed to delineate the explicit clinical and prognostic drivers of surgical decision-making to optimize treatment allocation in ICH.

**Trial Registration**: Clinicaltrials.gov Identifier: ATACH-2, NCT01176565; ERICH, NCT01202864

{{< pagebreak >}}

## Introduction

Intracerebral hemorrhage (ICH) remains a challenging disease, marked by its sudden onset, morbidity, and mortality. [@hemphillICHScoreSimple2001; @shethSpontaneousIntracerebralHemorrhage2022] While management of ischemic stroke has advanced considerably over the last decade, the acute management of ICH remains limited to blood pressure control, coagulopathy reversal, and perhaps surgical evacuation of hemorrhage. [@pradillaTrialEarlyMinimally2024; @bakoMinimallyInvasiveSurgery2023] These efforts have focused on reducing mortality and preventing additional morbidity due to secondary injury.

Clinical presentations for patients with stroke are varied based on the hemisphere affected, where hemispheric dominance dictates cognitive processing, attention, language, and emotional responses. Consequently, when confronted with ICH, the specific hemisphere affected can profoundly influence not only the clinical presentation but also management decisions and prognostic outcomes.

Previous research involving patients with ischemic stroke provide evidence that those with right hemispheric injury present later to the hospital and may not be treated as frequently with thrombolytic therapy. [@dileggeImpactLesionSide2005; @foerchDifferenceRecognitionRight2005; @portegiesLeftsidedStrokesAre2015; @liStrokeLateralizationLarge2021] Some practitioners hesitate to offer decompressive craniectomy in patients with left dominant-hemispheric ischemic infarcts despite literature demonstrating that long-term functional outcomes and health-related quality of life (HRQoL) are equivalent if not better for patients with left hemispheric injuries. [@deb-chatterjiSideMattersDifferences2022; @leggeNeglectingDifference2006; @beuscherFunctionalLongTermOutcome2017; @satoHigherMortalityPatients2015]

Amidst an increasing enthusiasm for surgical management of ICH [@bakoMinimallyInvasiveSurgery2023; @pradillaTrialEarlyMinimally2024; @beckDecompressiveCraniectomyBest2024], it is important to understand whether hemispheric lateralization is associated with differences in surgical decision making, as has been observed for ischemic stroke. We aimed to evaluate the association of hemispheric lateralization of ICH with the likelihood of receiving neurosurgical intervention as well as with functional outcomes.

## Methods

### Study Design and Participants

We conducted a post-hoc analysis using data from two prospective studies: the Antihypertensive Treatment of Acute Cerebral Hemorrhage II (ATACH-2) trial and the Ethnic/Racial Variations of Intracerebral Hemorrhage (ERICH) study. ATACH-2 was a large, multicenter, international, open-label, blinded-endpoint randomized controlled trial that enrolled 1000 patients with ICH and randomized them to either intensive blood pressure control (110-139 mm Hg) or standard blood pressure control (140-179 mm Hg). ERICH was a large, multicenter, prospective observational study conducted patients in the United States on patients admitted with ICH, with particular emphasis on recruitment of a racially and ethnically diverse cohort. The protocols and results of both studies have been previously published elsewhere [@qureshiIntensiveBloodPressureLowering2016; @wooEthnicRacialVariations2013].

We included all patients enrolled in these two studies with supratentorial location of ICH. We excluded patients with cerebellar or brainstem ICH as well as any patient with primary intraventricular hemorrhage (IVH).

The study was approved by the institutional review board of the Wake Forest School of Medicine (IRB00098198).

### Exposures and Outcomes

Our exposure was hemispheric laterality (left versus right). Our primary outcome was neurosurgical intervention as a binary variable defined as receiving a hematoma evacuation, decompressive craniectomy, or other craniotomy.

To explore differences in other aspects of aggressive care, we defined several secondary outcomes: receipt of external ventricular drain (EVD), receipt of tracheostomy, days of mechanical ventilation, receipt of do-not-resuscitate (DNR) order at any point in the hospitalization, withdrawal of life-sustaining therapy (WLST) at any point in the hospitalization, and early WLST (defined as \< 72 from symptom onset).

Our primary functional outcome was modified Rankin score (mRS) at 90 days analyzed via ordinal regression. We also examined quality of life as assessed by the European Quality of Life-5 dimensions (EQ-5D) questionnaire.

### Statistical Analysis

Associations between hemispheric laterality and clinical outcomes were evaluated using hierarchical Bayesian regression models, adjusting for the covariates listed below with a random intercept applied to the enrolling study cohort.

We employed Bayesian statistical methods for this analysis due to several advantages over traditional frequentist approaches. Bayesian methods allow for the incorporation of prior knowledge or beliefs into the analysis, which can be particularly useful in observational studies where prior information about relationships between variables may exist. This is achieved through the use of prior distributions, which are combined with the observed data to produce posterior distributions representing updated beliefs about the parameters of interest. Furthermore, Bayesian methods provide a natural framework for quantifying uncertainty through credible intervals (CrI), which offer a more intuitive interpretation than frequentist confidence intervals. Finally, Bayesian modeling facilitates the estimation of posterior probabilities, allowing for direct probabilistic statements about hypotheses, such as the probability that an effect size (e.g., odds ratio [OR] or incidence rate ratio [IRR]) is greater than 1, which is often of greater clinical relevance than a p-value. [@dogucuBayesRulesIntroduction; @mcelreathStatisticalRethinkingBayesian2020] This report adheres to the Bayesian Analysis Reporting Guidelines. [@kruschkeBayesianAnalysisReporting2021] 

Binary, ordinal, and count-based outcomes were modeled using hierarchical Bayesian logistic, ordinal logistic, and zero-inflated negative binomial regressions, respectively. The European Quality of Life Visual Analog Scale (EQ-VAS) was scaled to a proportion and analyzed via hierarchical Bayesian zero-one inflated beta (ZOIB) regression. Because our adjusted models included an interaction term between hemispheric laterality and ICH location, the raw main-effect coefficients for laterality are conditional on the reference location. To calculate the overall average effect of laterality across the entire cohort, we applied marginal standardization (G-computation) to all models. This allowed us to estimate average marginal effects while averaging over the empirical distribution of all covariates and the random intercept.

To account for potential clustering of clinical decision-making by hospital, we ideally would have included a random intercept for the study site in all models. However, site-level identifiers were not available for the ERICH cohort. To assess the potential impact of omitting site-level clustering, we performed a sensitivity analysis restricted to the ATACH-2 cohort, where site identifiers were available. We fitted the primary model for neurosurgical intervention with and without a random intercept for hospital site and compared the posterior estimates for the association with laterality.

All analyses were performed in R version 4.4.2 (R Foundation for Statistical Computing, Vienna, Austria). All packages used for the analysis are listed in the Supplementary Appendix.

#### Covariate Selection

We constructed directed acyclic graphs (DAGs) to model our causal assumptions and to guide covariate selection for each estimand. [@debertinSynthesizingSubjectmatterExpertise2024; @tennantUseDirectedAcyclic2021; @hernanSpecifyingTargetTrial2016] For neurosurgical intervention and other acute care outcomes, the target estimand was the total effect of hemispheric laterality on treatment decisions. Models were adjusted for baseline confounders that independently influence both the clinical presentation associated with hemispheric laterality and the likelihood of intervention: age, admission Glasgow Coma Scale (GCS) score, ICH location (categorized as lobar, thalamic, or basal ganglia), baseline ICH volume, and presence of intraventricular hemorrhage (IVH).

For functional outcomes (mRS and EQ-5D), the target estimand was the total effect of hemispheric laterality, encompassing all causal pathways including those mediated through neurosurgical intervention and other post-baseline care decisions. These models were adjusted for the same baseline confounders but did not condition on neurosurgical intervention, as it lies on the causal pathway from laterality to functional outcome; adjusting for a mediator would block the clinically relevant pathway and risk introducing collider stratification bias. [@tennantUseDirectedAcyclic2021] The DAGs and further details of the covariate selection rationale are presented in the Supplementary Appendix.

#### Prior Distribution Specification

To assess the robustness of our findings, we applied four sets of prior distributions for the primary exposure across all models. Because the models utilize different link functions depending on the outcome distribution (e.g., logit for binary/ordinal outcomes, log for count outcomes), these priors correspond to log-odds ratios (OR), log-incidence rate ratios (IRR), or log-mean differences. 

The four prior sets were: (1) a neutral, weakly informative prior centered on a null effect (Normal(0, 0.5)); (2) an informative left-hemisphere prior favoring outcomes in left-hemispheric patients (Normal(-0.22, 0.175)); (3) an informative right-hemisphere prior favoring outcomes in right-hemispheric patients (Normal(0.18, 0.175)); and (4) a vague, diffuse prior (Normal(0, 5)) that provides minimal information to the posterior, approximating the assumptions of a frequentist analysis. Further rationale and empirical evaluations of these priors are detailed in the Supplementary Appendix.

#### Computation and Description of the Posterior Distribution

For the effect of hemispheric laterality, we report the following metrics derived from the marginalized posterior distributions: 1) the median posterior effect size with 95% credible intervals (CrI); 2) the probability of any directional difference; 3) the probability of a substantial difference defined a priori; and 4) the percentage of the posterior distribution falling outside a predefined region of practical equivalence (ROPE). 

Effect sizes are reported as marginal odds ratios (OR) for binary and ordinal models, marginal incidence rate ratios (IRR) for the count component of the negative binomial models, and average marginal expected mean differences on the original 0–100 scale for the EQ-VAS (derived via G-computation). We defined a substantial difference as an odds ratio (OR) > 1.2 (or a mean difference of 5 points for the Euro VAS). In the context of invasive procedures such as hemicraniectomy or hematoma evacuation, we posited that a 20% relative shift in treatment odds represents a minimum threshold that would meaningfully impact clinical and shared decision-making. The full rationale for these thresholds and the ROPE bounds for ORs and IRRs are detailed in the Supplementary Appendix.

#### Missing Data Treatment and Sensitivity Analyses

Baseline covariates used in model adjustment (e.g., age, GCS, ICH volume) had minimal missingness (< 1%). Missingness for 90-day functional outcomes was more substantial but was balanced across laterality groups. Detailed missingness rates, assessments supporting the Missing at Random (MAR) assumption, and imputation diagnostics are reported in the Supplementary Appendix.

Our primary analysis for 90-day functional outcomes used multiple imputation by chained equations (MICE) to address missing data, generating 20 imputed datasets using the mice package. Imputation models included all analysis covariates and all primary and secondary outcomes. Including all functional outcomes (e.g., mRS and EuroQOL domains) in the imputation equations allowed them to serve as highly informative auxiliary variables for one another, optimizing the imputation of missing outcome data. Predictive mean matching was used for continuous and ordinal variables, logistic regression for binary variables, and polytomous regression for nominal variables. Each Bayesian model was fitted separately to each imputed dataset, and posterior draws were pooled across imputations to produce the reported estimates and CrI.

In addition to the missing data analysis, we analyzed our models using the four sets of prior distributions described above. To evaluate potential effect modification on acute care decisions, we fitted models containing formal interaction terms between hemispheric laterality and ICH location, and laterality and study source. Because missingness in baseline covariates and acute surgical outcomes was negligible, these exploratory subgroup interaction models—as well as a prespecified sensitivity analysis for all outcomes—were fitted to complete-case data (excluding patients missing the specific outcome being modeled) for computational efficiency. Subgroup-specific aORs and interaction Ratio of Odds Ratios (ROR) with 95% CrI were computed to evaluate heterogeneity.

### Data Availability Statement

The ATACH-2 (PI: Adnan I. Qureshi, MD, NCT01176565) and ERICH (PI: Daniel Woo, MD, MS, NCT01202864) data and analyses presented in this manuscript are based on the research files downloaded from the Archived Clinical Research web site of the National Institute of Neurologic Disease and Stroke (NINDS). The datasets can be obtained after approval of a written proposal as outlined on the NINDS website. As outlined in the NINDS Data Request Agreement, the analysis code is publicly available at <https://github.com/nikhilpatelmd/ich-laterality>.

## Results

### Baseline Characteristics

A total of `r sample_size %>% pull(n)` patients were included in the analysis, of which `r left_hemisphere_count %>% pull(n)` (`r percent(left_hemisphere_count %>% pull(n) / sample_size %>% pull(n), digits = 1)`) had left hemispheric ICH (Table 1). They had higher National Institutes of Health Stroke Scale (NIHSS) scores, lower GCS scores, and lower rates of hypertension as compared to those with right hemispheric ICH. There were `r erich_count %>% pull(n)` (`r percent(erich_count %>% pull(n) / sample_size %>% pull(n), digits = 1)`) patients in the cohort from ERICH and `r atach_count %>% pull(n)` (`r percent(atach_count %>% pull(n) / sample_size %>% pull(n), digits = 1)`) from ATACH-2.

### Hemispheric Laterality and Neurosurgical Intervention


Associations between hemispheric laterality and primary and secondary outcomes are shown in Table 2. `r tibble2 %>% filter(Outcome == "Neurosurgical Intervention") %>% pull(Left)` patients with left hemispheric ICH underwent neurosurgical intervention, compared to `r tibble2 %>% filter(Outcome == "Neurosurgical Intervention") %>% pull(Right)` with right hemispheric ICH (aOR `r odds_ratio_surgery`, 95% CrI `r lower_bound_surg`–`r upper_bound_surg`). The posterior probability that the aOR exceeded 1 was `r tibble2 %>% filter(Outcome == "Neurosurgical Intervention") %>% pull(or_1)`, and the posterior probability of a substantial difference (aOR > 1.2) was `r tibble2 %>% filter(Outcome == "Neurosurgical Intervention") %>% pull(or_1.2)`. Fewer than `r tibble2 %>% filter(Outcome == "Neurosurgical Intervention") %>% pull(rope)` of the posterior distribution fell within the ROPE.

In exploratory subgroup analyses, the association between right hemispheric ICH and neurosurgical intervention remained directionally consistent across all evaluated ICH locations and source studies. Formal interaction analyses yielded wide credible intervals for the Ratio of Odds Ratios (ROR), providing insufficient evidence to conclude that the laterality effect differed meaningfully across these subgroups (Supplemental Table 38).

Figure 1 displays the covariate-adjusted marginal predicted probabilities of neurosurgical intervention and the prior sensitivity analysis. Panel A shows the posterior distributions of covariate-adjusted absolute predicted probabilities of neurosurgical intervention for left and right hemispheric ICH under the neutral prior (left hemisphere: 4.86%, 95% CrI 3.98–5.84%; right hemisphere: 8.04%, 95% CrI 6.93–9.23%). Panel B shows the posterior distributions of the absolute probability difference (Right − Left) across all four prior specifications. Under the neutral prior, the median probability difference was 3.18 percentage points (95% CrI 1.75–4.62), with P(Right > Left) exceeding 99.9%. Corresponding estimates were 2.21 percentage points (95% CrI 0.97–3.45) under the left-hemisphere prior, 2.84 percentage points (95% CrI 1.60–4.09) under the right-hemisphere prior, and 3.38 percentage points (95% CrI 1.90–4.89) under the diffuse prior, with P(Right > Left) exceeding 99.9% under all specifications. Further details are available in Supplemental Tables 39-41.

In a sensitivity analysis restricted to the ATACH-2 cohort, a random intercept for the enrolling hospital was included to account for site-level clustering. The base model yielded an adjusted odds ratio of 2.23 (95% CrI: 1.31–3.76). With the inclusion of the site-level random effect, the adjusted odds ratio was 1.86 (95% CrI: 1.10–3.16). For this hierarchical model, the posterior probability of any directional effect (aOR > 1) was 99%, the posterior probability of an aOR > 1.2 was 95%, and 1% of the posterior distribution fell within the region of practical equivalence (Supplementary Table 46)

### Hemispheric Laterality and Secondary Outcomes

A total of `r tibble2 %>% filter(Outcome == "External Ventricular Drain") %>% pull(Left)` with left hemispheric ICH had an EVD placed as compared to `r tibble2 %>% filter(Outcome == "External Ventricular Drain") %>% pull(Right)` with right hemispheric ICH (aOR `r odds_ratio_evd`, 95% CrI `r lower_bound_evd`–`r upper_bound_evd`). The posterior probability that the aOR exceeded 
1 was `r tibble2 %>% filter(Outcome == "External Ventricular Drain") %>% pull(or_1)`, and the posterior probability of a substantial difference (aOR > 1.2) was `r tibble2 %>% filter(Outcome == "External Ventricular Drain") %>% pull(or_1.2)`. `r tibble2 %>% filter(Outcome == "External Ventricular Drain") %>% pull(rope)` of the posterior distribution fell within the ROPE.

Median days on mechanical ventilation were `r tibble2 %>% filter(Outcome == "Days of Mechanical Ventilation") %>% pull(Left)` and `r tibble2 %>% filter(Outcome == "Days of Mechanical Ventilation") %>% pull(Right)` for left and right hemispheric ICH, respectively (IRR `r irr_mv`, 95% CrI `r lower_bound_mv`–`r upper_bound_mv`). The posterior probability that the IRR exceeded 1 was `r tibble2 %>% filter(Outcome == "Days of Mechanical Ventilation") %>% pull(or_1)`, and `r tibble2 %>% filter(Outcome == "Days of Mechanical Ventilation") %>% pull(rope)` of the posterior distribution fell within the ROPE.

Rates of tracheostomy were similar between groups (`r tibble2 %>% filter(Outcome == "Tracheostomy") %>% pull(Left)` left vs. `r tibble2 %>% filter(Outcome == "Tracheostomy") %>% pull(Right)` right; aOR `r tibble2 %>% filter(Outcome == "Tracheostomy") %>% pull(or_ci)`), with a posterior probability that the aOR exceeded 1 of `r tibble2 %>% filter(Outcome == "Tracheostomy") %>% pull(or_1)` and `r tibble2 %>% filter(Outcome == "Tracheostomy") %>% pull(rope)` of the posterior distribution within the ROPE. Rates of WLST (`r tibble2 %>% filter(Outcome == "Withdrawal of Life-Sustaining Therapy") %>% pull(Left)` left vs. `r tibble2 %>% filter(Outcome == "Withdrawal of Life-Sustaining Therapy") %>% pull(Right)` right; aOR `r tibble2 %>% filter(Outcome == "Withdrawal of Life-Sustaining Therapy") %>% pull(or_ci)`), early WLST (`r tibble2 %>% filter(Outcome == "Early Withdrawal of Life-Sustaining Therapy") %>% pull(Left)` left vs. `r tibble2 %>% filter(Outcome == "Early Withdrawal of Life-Sustaining Therapy") %>% pull(Right)` right; aOR `r tibble2 %>% filter(Outcome == "Early Withdrawal of Life-Sustaining Therapy") %>% pull(or_ci)`), and DNR orders (`r tibble2 %>% filter(Outcome == "DNR Order") %>% pull(Left)` left vs. `r tibble2 %>% filter(Outcome == "DNR Order") %>% pull(Right)` right; aOR `r tibble2 %>% filter(Outcome == "DNR Order") %>% pull(or_ci)`) were also similar across hemispheres, with P(aOR > 1) of `r tibble2 %>% filter(Outcome == "Withdrawal of Life-Sustaining Therapy") %>% pull(or_1)`, `r tibble2 %>% filter(Outcome == "Early Withdrawal of Life-Sustaining Therapy") %>% pull(or_1)`, and `r tibble2 %>% filter(Outcome == "DNR Order") %>% pull(or_1)`, respectively.

### Hemispheric Laterality and Functional Outcomes

```{r functional-outcomes}
# mRS
or_ci_values_mrs <- tibble3 %>%
  filter(outcome == "Modified Rankin Score") %>%
  pull(est_label)

or_ci_split_mrs <- str_split(or_ci_values_mrs, " \\(| - ", simplify = TRUE)

odds_ratio_mrs <- or_ci_split_mrs[[1]]
lower_bound_mrs <- or_ci_split_mrs[[2]]
upper_bound_mrs <- str_remove(or_ci_split_mrs[[3]], "\\)")

# EuroQOL - Mobility
or_ci_values_mob <- tibble3 %>%
  filter(outcome == "EuroQOL - Mobility") %>%
  pull(est_label)

or_ci_split_mob <- str_split(or_ci_values_mob, " \\(| - ", simplify = TRUE)

odds_ratio_mob <- or_ci_split_mob[[1]]
lower_bound_mob <- or_ci_split_mob[[2]]
upper_bound_mob <- str_remove(or_ci_split_mob[[3]], "\\)")

# Euro VAS
or_ci_values_vas <- tibble3 %>%
  filter(outcome == "Euro VAS") %>%
  pull(est_label)

or_ci_split_vas <- str_split(or_ci_values_vas, " \\(| - ", simplify = TRUE)

mean_diff_vas <- or_ci_split_vas[[1]]
lower_bound_vas <- or_ci_split_vas[[2]]
upper_bound_vas <- str_remove(or_ci_split_vas[[3]], "\\)")
```

Associations between hemispheric laterality and functional outcomes are shown in Table 3. The aOR for a worse mRS score at 90 days for right hemispheric ICH compared to left hemispheric ICH was `r odds_ratio_mrs` (95% CrI `r lower_bound_mrs`–`r upper_bound_mrs`), with a posterior probability that the aOR exceeded 1 of `r tibble3 %>% filter(outcome == "Modified Rankin Score") %>% pull(prob_diff)` and `r tibble3 %>% filter(outcome == "Modified Rankin Score") %>% pull(prob_rope)` of the posterior distribution within the ROPE. Figure 3 displays the covariate-adjusted absolute marginal predicted cumulative distributions of mRS at 90 days for left and right hemispheric ICH under the neutral prior.

For Mobility and Self-Care, the posterior probability that the aOR exceeded 1 was `r tibble3 %>% filter(outcome == "EuroQOL - Mobility") %>% pull(prob_diff)` for both domains (aOR `r odds_ratio_mob`, 95% CrI `r lower_bound_mob`–`r upper_bound_mob` for Mobility; aOR `r tibble3 %>% filter(outcome == "EuroQOL - Self-Care") %>% pull(est_label)` for Self-Care), with a posterior probability of a substantial difference (aOR > 1.2) of `r tibble3 %>% filter(outcome == "EuroQOL - Mobility") %>% pull(prob_sub)` for both, and `r tibble3 %>% filter(outcome == "EuroQOL - Mobility") %>% pull(prob_rope)` and `r tibble3 %>% filter(outcome == "EuroQOL - Self-Care") %>% pull(prob_rope)` of the posterior distributions within the ROPE, respectively. For Pain/Discomfort, the aOR was `r tibble3 %>% filter(outcome == "EuroQOL - Pain/Discomfort") %>% pull(est_label)` with P(aOR > 1) of `r tibble3 %>% filter(outcome == "EuroQOL - Pain/Discomfort") %>% pull(prob_diff)` and `r tibble3 %>% filter(outcome == "EuroQOL - Pain/Discomfort") %>% pull(prob_rope)` of the posterior distribution within the ROPE. For Anxiety/Depression, the aOR was `r tibble3 %>% filter(outcome == "EuroQOL - Anxiety/Depression") %>% pull(est_label)` with P(aOR > 1) of `r tibble3 %>% filter(outcome == "EuroQOL - Anxiety/Depression") %>% pull(prob_diff)` and `r tibble3 %>% filter(outcome == "EuroQOL - Anxiety/Depression") %>% pull(prob_rope)` of the posterior distribution within the ROPE. For Usual Activities, the aOR was `r tibble3 %>% filter(outcome == "EuroQOL - Usual Activities") %>% pull(est_label)` with P(aOR > 1) of `r tibble3 %>% filter(outcome == "EuroQOL - Usual Activities") %>% pull(prob_diff)` and `r tibble3 %>% filter(outcome == "EuroQOL - Usual Activities") %>% pull(prob_rope)` of the posterior distribution within the ROPE. For the Euro VAS, the estimated mean difference was `r mean_diff_vas` points (95% CrI `r lower_bound_vas`–`r upper_bound_vas`), with a posterior probability of `r tibble3 %>% filter(outcome == "Euro VAS") %>% pull(prob_diff)` that right hemispheric ICH was associated with a lower score, `r tibble3 %>% filter(outcome == "Euro VAS") %>% pull(prob_rope)` of the posterior distribution within the ROPE, and a posterior probability of a substantial difference (mean difference < −5 points) of `r tibble3 %>% filter(outcome == "Euro VAS") %>% pull(prob_sub)`. 

Results from sensitivity analyses using alternative priors and complete case analysis are presented in Supplementary Tables 42-45 and are consistent with these results.

## Discussion

In this pooled analysis of r sample_size patients with ICH, we observed an association between right hemispheric ICH and a higher likelihood of receiving neurosurgical intervention compared to left hemispheric ICH. Specifically, patients with right hemispheric ICH had `r odds_ratio_surgery` times the adjusted odds of receiving a neurosurgical intervention and `r odds_ratio_evd` times the adjusted odds of receiving an EVD. This association was directionally consistent across ICH locations, study cohorts, and prior specifications, though incorporating a hospital-level random intercept to account for site-level clustering modestly attenuated the effect. Despite less aggressive surgical management, patients with left hemispheric ICH demonstrated similar 90-day functional outcomes and potentially better quality of life measures. This equivalence aligns with previous literature in both ICH and ischemic stroke, where a clear association between hemispheric laterality and functional outcomes has often been difficult to establish.

The exact mechanisms driving these associations remain unclear, and the observational nature of the data precludes causal inferences. In both the ATACH-2 and ERICH studies, neurosurgical candidacy was not governed by strict trial protocols but was left entirely to clinician discretion. Furthermore, ATACH-2 excluded patients who required immediate, life-saving hematoma evacuation upon presentation. Therefore, the observed disparity in our cohort likely reflects a complex interplay of clinical triage, unmeasured anatomical severity, and shared decision-making rather than isolated treatment bias.

One plausible explanation for this disparity involves language-related prognostic framing. Left-hemispheric ICH is frequently associated with aphasia, a deficit that clinicians, patients, and families historically weigh heavily when anticipating long-term quality of life. [@neugebauerDESTINYSAttitudesPhysicians2014; @neugebauerAttitudesPatientsRelatives2017; @neugebauerAttitudesNursesDisability2019] Prognostic framing that emphasizes the risk of severe aphasia may heavily influence surrogate decision-makers during early goals-of-care discussions, potentially driving lower intervention rates. This dynamic is consistent with the well-characterized self-fulfilling prophecy in neurocritical care. Conversely, severe neglect—which commonly complicates right-hemispheric lesions and profoundly impairs mobility and self-care—may be less frequently emphasized as a barrier to surgical candidacy, despite its distinct impact on caregiver burden and long-term disability.

This study has several important limitations. First, we lacked granular data on critical confounders such as specific mass effect trajectories, degree of midline shift, hydrocephalus, and explicit family preferences. These unmeasured clinical nuances are primary drivers of surgical decision-making and could independently influence both perceived candidacy and outcomes. Second, we relied on hemispheric laterality as an imperfect proxy for true language dominance; because handedness data were unavailable, the expected 5–10% of patients with atypical dominance likely introduced non-differential misclassification, biasing our estimates toward the null. Third, absolute surgical event rates in this cohort were relatively low. Consequently, it is unclear if these associations are transportable to contemporary, unselected ICH populations or to patients with moderate-to-severe ICH who possess immediate indications for emergent surgery. Finally, attitudes toward disability and the frameworks used for shared decision-making carry significant cultural nuance, and our findings—derived primarily from cohorts in the United States—may not generalize across diverse global and socioeconomic settings.

With recent randomized trials demonstrating the benefit of hematoma evacuation and decompressive craniectomy in appropriately selected ICH patients [@pradillaTrialEarlyMinimally2024; @bakoMinimallyInvasiveSurgery2023], understanding the specific drivers of discretionary surgical decisions is increasingly critical. Future research should prospectively investigate the precise clinical criteria used to determine surgical candidacy, while also exploring the details of prognostic framing and shared decision-making across varying cultural contexts to ensure equitable and evidence-based neurosurgical care.

## Conclusions

In this post-hoc observational analysis, right hemispheric ICH was associated with a higher likelihood of neurosurgical intervention compared to left hemispheric ICH. However, 90-day functional outcomes were similar across hemispheres. While these findings highlight a potential laterality-driven disparity in treatment paradigms, residual confounding from unmeasured anatomical severity and family preferences cannot be excluded. Future prospective studies are needed to delineate the explicit clinical and prognostic drivers of surgical decision-making to optimize treatment allocation in ICH.

{{< pagebreak >}}

## Acknowledgments

**Funding/support**: The ATACH-2 clinical trial and ERICH prospective observational study were funded by the National Institute of Neurological Disorders and Stroke (NINDS). 

**Role of the Funder/Sponsor**: The study was supported in part by grant 5U01NS062091, NINDS and grant 5U01NS069763, NINDS. The sponsor had no role in the design and conduct of the current study.

**Conflict of Interest Disclosures**: Dr. Magun discloses being a shareholder in Inovio Pharmaceuticals and StageZero Life Sciences.

**Additional Contributions**: We thank all investigators for their efforts in conducting the ATACH-2 trial and ERICH study and thank all study participants for enrolling.

{{< pagebreak >}}

## References

::: {#refs}
:::

{{< pagebreak >}}

## Tables

```{r}
#| echo: false
#| output: true
#| tbl-cap: "Demographic and Clinical Characteristics of Patients with Left and Right Hemispheric Intracerebral Hemorrhage at Baseline"
#| label: tbl-1
table_1
```

{{< pagebreak >}}


```{r}
#| echo: false
#| output: true
#| tbl-cap: "Association of Hemispheric Laterality with Primary and Secondary Outcomes"
#| label: tbl-2
table_2_neutral
```

{{< pagebreak >}}


```{r}
#| echo: false
#| output: true
#| tbl-cap: "Association of Hemispheric Laterality with Functional Outcomes"
#| label: tbl-3
table_3_neutral
```

{{< pagebreak >}}

## Figures

![Prior Sensitivity Analysis for Neurosurgical Intervention](../figures/figure_sensitivity_neurosurgery_evac.pdf){#fig-1}

![modified Rankin Score at 90 days Stratified by ICH Laterality](../figures/mrs/figure_mrs_main_neutral_adjusted.pdf){#fig-2}
