#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_neurosurgery))
with_dir(here(), tar_read(prior_visual_neurosurgery_neutral))
    corresponding: true
  - name: Matt Jaffa
    affiliation: Department of Neurology, University of Connecticut School of Medicine
  - name: Rahul Karamchandani
    affiliation: Department of Neurology, Wake Forest University School of Medicine
  - name: Andrew Asimos
    affiliation: Department of Emergency Medicine, Wake Forest University School of Medicine
  - name: Jonathan Parish
    affiliation: Carolina Neurosurgery & Spine Associates
  - name: Nicholas Allen Morris
    affiliation: Department of Neurology, University of Maryland School of Medicine
date: last-modified
abstract: |
bibliography: references.bib
csl: american-heart-association.csl
table-of-contents: true
toc-title: Contents
---

```{r setup}
#| echo: false
#| output: false

#| echo: false
#| fig-height: 12

library(targets)
library(tidyverse)
library(withr)
library(here)
library(ggdag)
library(modelsummary)
library(marginaleffects)
library(brms)
library(tidybayes)
library(bayesplot)
library(glue)
library(ggdist)
library(patchwork)
library(tinytable)
library(gt)


# plot stuff
theme_ich <- function(base_family = "Figtree",
                      base_size = 10) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(lineheight = 2),
      plot.title = element_text(size = rel(1.3), margin = margin(12, 0, 8, 0)),
      plot.subtitle = element_text(size = rel(1.1), margin = margin(4, 0, 0, 0)),
      axis.text.y = element_text(size = rel(0.8)),
      axis.title.y = element_text(size = 12, margin = margin(0, 4, 0, 0)),
      axis.text.x = element_text(size = rel(2)),
      axis.title.x = element_blank(),
      legend.position = "bottom",
      legend.justification = 1,
      panel.grid = element_line(colour = "#F3F4F5"),
      plot.caption = element_text(size = rel(0.8), margin = margin(8, 0, 0, 0)),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25, "cm")
    )
}

# load data and models
with_dir(here(), tar_load(ich_aggressive))

```

# Details of Statistical Analysis

## Covariate Selection

```{r dags}
#| echo: false
#| warning: false
#| fig-height: 8

with_dir(here(), tar_load(dag_aggressive))

status_colors <- c(exposure = "#0074D9", outcome = "#FF4136", latent = "grey50")

dag_aggressive |>
  tidy_dagitty() |>
  node_status() |>
  ggdag() +
  geom_dag_edges() +
  geom_dag_point(aes(color = status)) +
  geom_dag_label_repel(aes(label = label, fill = status),
    seed = 1234,
    color = "white", fontface = "bold"
  ) +
  scale_color_manual(values = status_colors, na.value = "grey20") +
  scale_fill_manual(values = status_colors, na.value = "grey20") +
  guides(color = "none", fill = "none") +
  theme_dag()

```

## Prior Distribution Definitions

```{r prior definitions}
#| echo: false
#| output: false

neutral_mean <- 0
neutral_sd <- 0.35
neutral_lower_OR <- exp(neutral_mean - (1.96 * neutral_sd))
neutral_upper_OR <- exp(neutral_mean + (1.96 * neutral_sd))

left_mean <- -0.22
left_sd <- 0.175
left_lower_OR <- exp(left_mean - (1.96 * left_sd))
left_upper_OR <- exp(left_mean + (1.96 * left_sd))

right_mean <- 0.18
right_sd <- 0.175
right_lower_OR <- exp(right_mean - (1.96 * right_sd))
right_upper_OR <- exp(right_mean + (1.96 * right_sd))

```

In defining prior distributions, we generally followed suggestions from previous literature. [@zampieriUsingBayesianMethods2021; @kruschkeBayesianAnalysisReporting2021; @gelmanBayesianWorkflow2020] Specifically, we defined 4 sets of priors for each outcome:

**Neutral prior**: mean of **`r neutral_mean`** and standard deviation of **`r neutral_sd`**, centering around an OR of **`r exp(neutral_mean)`** and 95% of the the probability mass for the OR between **`r round(neutral_lower_OR, digits = 2)`** to **`r round(neutral_upper_OR, digits = 2)`**.
    
**Left hemisphere prior**: mean of **`r left_mean`** and standard deviation of **`r left_sd`**, centering around an OR of **`r round(exp(left_mean), digits = 2)`** and 95% of the the probability mass for the OR between **`r round(left_lower_OR, digits = 2)`** to **`r round(left_upper_OR, digits = 2)`**.

**Right hemisphere prior**: mean of **`r right_mean`** and standard deviation of **`r right_sd`**, centering around an OR of **`r round(exp(right_mean), digits =2)`** and 95% of the the probability mass for the OR between **`r round(right_lower_OR, digits = 2)`** to **`r round(right_upper_OR, digits =2)`**.

**Flat prior**: most mimics a frequentist analysis
    
```{r prior graphical summary}
#| echo: false
#| fig-height: 6

# ggplot(data = tibble(x = -2:2), aes(x = x)) +
#   geom_function(fun = dnorm, args = list(mean = -20, sd = 0.5), linewidth = 1) +
#   # stat_function(fun = dnorm_limit, geom = "area", color = "orange", alpha = 0.2) +
#   geom_vline(xintercept = -0.22, linetype = "dashed", color = "orange", linewidth = 1) +
#   scale_x_continuous(breaks = seq(-2, 2, 0.5)) +
#   theme_ich()

```

### Prior Predictive Checks

#### Neurosurgical Intervention

```{r prior predictive check - neurosurgery}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_neurosurgery))

with_dir(here(), tar_read(prior_visual_neurosurgery_neutral))

```

#### Ventriculostomy

```{r prior predictive check - evd}
#| echo: false
#| warning: false


```

#### Days of Mechanical Ventilation

```{r prior predictive check - vent}
#| echo: false
#| warning: false


```


#### Tracheostomy

```{r prior predictive check - tracheostomy}
#| echo: false
#| warning: false


```


#### Comfort Care

```{r prior predictive check - comfort}
#| echo: false
#| warning: false


```


#### Early WLST

```{r prior predictive check - early wlst}
#| echo: false
#| warning: false


```

#### DNR

```{r prior predictive check - dnr}
#| echo: false
#| warning: false


```

## Posterior Simulation Diagnostics

```{r posterior diagnostics setup}
#| echo: false

posterior_diagnostics <- function(mod) {
  # plot code modified from Andrew Heiss (https://bayesf22-notebook.classes.andrewheiss.com/bayes-rules/13-chapter.html#simulating-the-posterior)

  trace_plot <- mod |>
    gather_draws(`^b_.*`, regex = TRUE) |>
    ggplot(aes(x = .iteration, y = .value, color = factor(.chain))) +
    geom_line(linewidth = 0.1) +
    scale_color_viridis_d(option = "rocket", end = 0.85) +
    facet_wrap(vars(.variable), scales = "free_y")

  trank_plot <- mod |>
    gather_draws(`^b_.*`, regex = TRUE) |>
    group_by(.variable) |>
    mutate(draw_rank = rank(.value)) |>
    ggplot(aes(x = draw_rank, color = factor(.chain))) +
    stat_bin(geom = "step", binwidth = 250, position = position_identity(), boundary = 0) +
    scale_color_viridis_d(option = "rocket", end = 0.85) +
    facet_wrap(vars(.variable), scales = "free_y") +
    theme(axis.text.y = element_blank(), axis.title.y = element_blank(), axis.ticks.y = element_blank())

  # pp_check <- pp_check(mod, ndraws = 500)
  # pp_check_bars <- pp_check(mod, ndraws = 500, type = "bars_grouped", group = "ich_laterality")

  return((trank_plot / trace_plot))
}


```

### MCMC Chain Convergence and Resolution

#### Neurosurgical Intervention

```{r posterior diagnostics neurosurgery}
#| echo: false
#| fig-height: 10

posterior_diagnostics(with_dir(here(), tar_read(m_posterior_neutral_neurosurgery)))

```

#### Ventriculostomy

```{r posterior diagnostics evd}
#| echo: false
#| fig-height: 10

posterior_diagnostics(with_dir(here(), tar_read(m_posterior_neutral_evd)))

```

#### Tracheostomy

```{r posterior diagnostics trach}
#| echo: false
#| fig-height: 10

posterior_diagnostics(with_dir(here(), tar_read(m_posterior_neutral_tracheostomy)))

```

#### Days of Mechanical Ventilation

```{r posterior diagnostics vent}
#| echo: false
#| fig-height: 10

posterior_diagnostics(with_dir(here(), tar_read(m_posterior_neutral_days_mechanical_ventilation)))

```

#### Comfort Care

```{r posterior diagnostics comfort}
#| echo: false
#| fig-height: 10

posterior_diagnostics(with_dir(here(), tar_read(m_posterior_neutral_comfort)))

```

#### Early WLST

```{r posterior diagnostics early wlst}
#| echo: false
#| fig-height: 10

posterior_diagnostics(with_dir(here(), tar_read(m_posterior_neutral_early_wlst)))

```

#### DNR

```{r posterior diagnostics dnr}
#| echo: false
#| fig-height: 10

posterior_diagnostics(with_dir(here(), tar_read(m_posterior_neutral_dnr_binary)))

```

### Posterior predictive checks

```{r posterior prediction function}
#| echo: false

post_pred <- function(mod) {
  pp_check <- pp_check(mod, ndraws = 500)

  pp_check_bars <- pp_check(mod, ndraws = 500, type = "bars_grouped", group = "ich_laterality")

  return(pp_check / pp_check_bars)
}

```

### Neurosurgical Intervention

```{r posterior predictions neurosurgery}
#| echo: false

post_pred(with_dir(here(), tar_read(m_posterior_neutral_neurosurgery)))

```

### Ventriculostomy

```{r posterior predictions evd}
#| echo: false

post_pred(with_dir(here(), tar_read(m_posterior_neutral_evd)))

```

### Tracheostomy

```{r posterior predictions trach}
#| echo: false

post_pred(with_dir(here(), tar_read(m_posterior_neutral_tracheostomy)))

```

### Days of Mechanical Ventilation

```{r posterior predictions vent}
#| echo: false

post_pred(with_dir(here(), tar_read(m_posterior_neutral_days_mechanical_ventilation)))

```

### Comfort Care

```{r posterior predictions comfort}
#| echo: false

post_pred(with_dir(here(), tar_read(m_posterior_neutral_comfort)))

```

### Early WLST

```{r posterior predictions early wlst}
#| echo: false

post_pred(with_dir(here(), tar_read(m_posterior_neutral_early_wlst)))

```

### DNR

```{r posterior predictions dnr}
#| echo: false

post_pred(with_dir(here(), tar_read(m_posterior_neutral_dnr_binary)))

```


## Sensitivity Analyses

### Different Priors

### Alternative Adjustment Sets

### Alternative Handling of Missing Data