---
title: Patients with Left Hemispheric Intracerebral Hemorrhage are Less Likely to Receive Neurosurgical Intervention
subtitle: Supplementary Appendix
authors:
  - name: Nikhil Patel
    orcid: 0000-0002-3193-8590
    affiliation: Department of Internal Medicine, Wake Forest University School of Medicine
    roles: data analysis, writing
    corresponding: true
  - name: Matt Jaffa
    affiliation: Department of Neurology, University of Connecticut School of Medicine
  - name: Rahul Karamchandani
    affiliation: Department of Neurology, Wake Forest University School of Medicine
  - name: Andrew Asimos
    affiliation: Department of Emergency Medicine, Wake Forest University School of Medicine
  - name: Jonathan Parish
    affiliation: Carolina Neurosurgery & Spine Associates
  - name: Nicholas Allen Morris
    affiliation: Department of Neurology, University of Maryland School of Medicine
date: last-modified
format:
  html:
    embed-resources: true
    echo: false
  docx: default
abstract: |
bibliography: references.bib
csl: american-heart-association.csl
table-of-contents: true
toc-title: Contents

---

```{r setup}
#| echo: false
#| output: false

#| echo: false
#| fig-height: 12

library(targets)
library(tidyverse)
library(withr)
library(here)
library(ggdag)
library(modelsummary)
library(marginaleffects)
library(brms)
library(tidybayes)
library(bayesplot)
library(glue)
library(ggdist)
library(patchwork)
library(tinytable)
library(gt)


# plot stuff
theme_ich <- function(base_family = "Figtree", base_size = 10) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(lineheight = 2),
      plot.title = element_text(size = rel(1.3), margin = margin(12, 0, 8, 0)),
      plot.subtitle = element_text(
        size = rel(1.1),
        margin = margin(4, 0, 0, 0)
      ),
      axis.text.y = element_text(size = rel(0.8)),
      axis.title.y = element_text(size = 12, margin = margin(0, 4, 0, 0)),
      axis.text.x = element_text(size = rel(2)),
      axis.title.x = element_blank(),
      legend.position = "bottom",
      legend.justification = 1,
      panel.grid = element_line(colour = "#F3F4F5"),
      plot.caption = element_text(size = rel(0.8), margin = margin(8, 0, 0, 0)),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25, "cm")
    )
}

# load data and models
with_dir(here(), tar_load(ich_aggressive))

```

# Details of Statistical Analysis

## Covariate Selection

As mentioned in the primary manuscript, we constructed directed acyclic graphs (DAGs) to facilitate covariate selection. DAGs are graphical models used to represent causal relationships between variables. They consist of nodes (representing variables) and directed arrows (representing causal influences). DAGs can help minimize several sources of bias that are inherent in many observational studies, such as confounding bias and selection bias. When constructed with subject-matter knowledge, they can provide a framework for making causal inferences from observational data. Specifically, they can help researchers with three key benefits:

1. **Identify minimal adjustment sets**: Determine the smallest set of variables to control for to estimate the causal effect of an exposure on an outcome. 

2. **Avoid over-adjustment**: Prevent controlling for variables that are mediators or colliders, which can bias the results.

3. **Clarify causal assumptions**: Make the underlying causal assumptions explicit and transparent [@hernanSpecifyingTargetTrial2016].

In constructing DAGs for this project, we used guidance and recommendations from previous literature [@tennantUseDirectedAcyclic2021; @sauerUseDirectedAcyclic2013; @barrettCausalInference2025]. 

### Primary and Secondary Outcomes

For our primary outcome model we specified the following DAG:

```{r neurosurgery dag}
#| echo: false
#| warning: false
#| fig-height: 6

library(dagitty)
library(ggdag, warn.conflicts = FALSE)
library(ggokabeito)

with_dir(here(), tar_load(dag_neurosurgery))

status_colors <- c(exposure = "#6f86d6", outcome = "#764ba2", latent = "grey50")

dag_neurosurgery |>
  tidy_dagitty() |>
  node_status() |>
  ggdag() +
  geom_dag_edges() +
  geom_dag_point(aes(color = status)) +
  geom_dag_label_repel(
    aes(label = label, fill = status),
    seed = 1234,
    color = "white",
    fontface = "bold"
  ) +
  scale_color_manual(values = status_colors, na.value = "grey20") +
  scale_fill_manual(values = status_colors, na.value = "grey20") +
  guides(color = "none", fill = "none") +
  theme_dag()

```

Before considering if covariate adjustment is indicated for our outcome of interest, we need to establish the difference between the total, direct, and indirect effect of hemispheric laterality on neurosurgical intervention and the effect with which we are interested. [@tennantUseDirectedAcyclic2021]

The **total effect** of hemispheric laterality (X) on neurosurgical intervention (Y) refers to the overall influence of X on Y, regardless of the mechanisms through which that influence operates. It is the effect we would observe if we simply compared outcomes between patients with left vs. right hemispheric ICH without controlling for any variables.

The **direct effect** of X on Y refers to the influence of X on Y that is not mediated by any other variables in the DAG. It is the effect of hemispheric laterality on neurosurgical intervention that remains after holding all mediating variables constant.

The **indirect effect** of X on Y refers to the influence of X on Y that is mediated by one or more intermediate variables. It is the effect of hemispheric laterality on neurosurgical intervention that operates through other variables in the DAG.

We were interested in the direct effect of hemispheric laterality on neurosurgical intervention. We hypothesized that this relationship could be mediated by other variables that are commonly used to determine if a patient will undergo neurosurgical intervention. To estimate this effect, we included to the following covariates in our statistical model for the primary outcome: age, admission GCS, ICH location, ICH volume, and presence of IVH.

Since we felt that there were not materially different causal structures at play with our secondary outcomes, we used a similar DAG for the following outcomes: receipt of external ventricular drain (EVD), receipt of tracheostomy, days of mechanical ventilation, receipt of do-not-resuscitate (DNR) order at any point in the hospitalization, withdrawal of life-sustaining therapy (WLST) at any point in the hospitalization, and early WLST (defined as < 72 from symptom onset). We included the same covariates in our statistical models for all secondary outcomes specified above.

### Functional Outcomes

We specified the following DAG when examining functional outcomes (e.g., modified Rankin Scale and EuroQOL measures):

```{r functional outcome dag}
#| echo: false
#| warning: false
#| fig-height: 8.5

with_dir(here(), tar_load(dag_outcomes))

dag_outcomes |>
  tidy_dagitty() |>
  node_status() |>
  ggdag() +
  geom_dag_edges() +
  geom_dag_point(aes(color = status)) +
  geom_dag_label_repel(
    aes(label = label, fill = status),
    seed = 1234,
    color = "white",
    fontface = "bold"
  ) +
  scale_color_manual(values = status_colors, na.value = "grey20") +
  scale_fill_manual(values = status_colors, na.value = "grey20") +
  guides(color = "none", fill = "none") +
  theme_dag()

```

We are again interested in the direct effect of hemispheric laterality on functional outcomes, independent of the influence of neurosurgical intervention. Therefore, our statistical models for functional outcomes were adjusted for age, admission GCS, ICH location, ICH volume, presence of IVH, and neurosurgical intervention. This approach allows us to isolate the prognostic significance of the affected hemisphere on functional outcomes.

## Priors

### Prior Distribution Definitions

```{r prior definitions}
#| echo: false
#| output: false

neutral_mean <- 0
neutral_sd <- 0.35
neutral_lower_OR <- exp(neutral_mean - (1.96 * neutral_sd))
neutral_upper_OR <- exp(neutral_mean + (1.96 * neutral_sd))

left_mean <- -0.22
left_sd <- 0.175
left_lower_OR <- exp(left_mean - (1.96 * left_sd))
left_upper_OR <- exp(left_mean + (1.96 * left_sd))

right_mean <- 0.18
right_sd <- 0.175
right_lower_OR <- exp(right_mean - (1.96 * right_sd))
right_upper_OR <- exp(right_mean + (1.96 * right_sd))

flat_mean <- 0
flat_sd <- 5
flat_lower_OR <- exp(flat_mean - (1.96 * flat_sd))
flat_upper_OR <- exp(flat_mean + (1.96 * flat_sd))

```

In defining prior distributions, we generally followed suggestions from previous literature. [@zampieriUsingBayesianMethods2021; @kruschkeBayesianAnalysisReporting2021; @gelmanBayesianWorkflow2020] Specifically, we defined 4 sets of priors for each outcome:

**Neutral prior**: This prior assumed that hemispheric lateralization did not significantly affect the primary outcome and represents the **null hypothesis**. It is centered around a mean of **`r neutral_mean`** and standard deviation of **`r neutral_sd`**, corresponding to OR of **`r exp(neutral_mean)`** and 95% of the probability mass for the OR between **`r round(neutral_lower_OR, digits = 2)`** to **`r round(neutral_upper_OR, digits = 2)`**.

**Left hemisphere prior**: This prior assumes that the primary and secondary outcomes are more likely with left hemisphere injury, centering around a mean of **`r left_mean`** and standard deviation of **`r left_sd`** which corresponds to an OR of **`r round(exp(left_mean), digits = 2)`** and 95% of the probability mass for the OR between **`r round(left_lower_OR, digits = 2)`** to **`r round(left_upper_OR, digits = 2)`**.

**Right hemisphere prior**: This prior assumes that the primary and secondary outcomes are more likely with right hemisphere injury, centering around a mean of **`r right_mean`** and standard deviation of **`r right_sd`** which corresponds to an OR of **`r round(exp(right_mean), digits =2)`** and 95% of the probability mass for the OR between **`r round(right_lower_OR, digits = 2)`** to **`r round(right_upper_OR, digits =2)`**.

**Flat prior**: This prior most mimics a frequentist analysis in that the prior distribution is so wide that it essentially contributes no information to the posterior model. It is centered around a mean of **`r flat_mean`** and standard deviation of standard deviation of **`r flat_sd`** which corresponds to an OR of **`r round(exp(flat_mean), digits =2)`** and 95% of the probability mass for the OR between **`r round(flat_lower_OR, digits = 2)`** to **`r round(flat_upper_OR, digits = 20)`**.

### Prior Predictive Checks

Prior predictive checks were performed to understand the implications of the prior distributions before analyzing the data and ensuring the prior distributions represented plausible information regarding each outcome.

#### Neurosurgical Intervention

```{r prior predictive check - neurosurgery}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_neurosurgery))

```

#### Ventriculostomy

```{r prior predictive check - evd}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_evd))


```

#### Tracheostomy

```{r prior predictive check - tracheostomy}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_tracheostomy))


```


#### Days of Mechanical Ventilation

```{r prior predictive check - vent}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_days_mechanical_ventilation))


```


#### Comfort Care

```{r prior predictive check - comfort}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_comfort))


```


#### Early WLST

```{r prior predictive check - early wlst}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_early_wlst))


```

#### DNR

```{r prior predictive check - dnr}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_dnr_binary))


```

#### mRS 90

```{r prior predictive check - mrs90}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_mrs_90))


```

#### EuroOOL Mobility 90

```{r prior predictive check - euro mobility}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_euro_mobility_90))


```

#### EuroOOL Self-Care 90

```{r prior predictive check - euro selfcare}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_euro_selfcare_90))


```

#### EuroOOL Usual Activities 90

```{r prior predictive check - euro usual}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_euro_usual_90))


```

#### EuroOOL Pain/Discomfort 90

```{r prior predictive check - euro pain}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_euro_pain_90))


```

#### EuroOOL Anxiety/Depression 90

```{r prior predictive check - euro anxiety}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_euro_anxiety_90))


```

#### EuroOOL VAS 90

```{r prior predictive check - vas}
#| echo: false
#| warning: false

with_dir(here(), tar_read(prior_check_euro_anxiety_90))


```

## Posterior Simulation Diagnostics

Posterior Simulation Checks
To ensure the reliability and validity of our Bayesian models, we performed several posterior simulation checks.

Chain Convergence: We assessed the convergence of the Markov Chain Monte Carlo (MCMC) chains by visually inspecting trace plots for each parameter. The trace plots showed that the chains were well-mixed and stationary, indicating that they had converged to a stable posterior distribution. Additionally, we calculated the R-hat statistic for all parameters, ensuring that all values were below 1.01, which confirms the convergence of the chains.

Posterior Predictive Checks: We conducted posterior predictive checks to evaluate the model's fit to the data. This involved simulating data from the posterior predictive distribution and comparing it to the observed data. The close correspondence between the simulated and observed data confirmed that our models provided a good fit to the underlying data structure.

### MCMC Chain Convergence and Resolution

#### Neurosurgical Intervention

```{r posterior diagnostics neurosurgery}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_neurosurgery))

```

#### Ventriculostomy

```{r posterior diagnostics evd}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_evd))

```

#### Tracheostomy

```{r posterior diagnostics trach}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_tracheostomy))

```

#### Days of Mechanical Ventilation

```{r posterior diagnostics vent}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_days_mechanical_ventilation))

```

#### Comfort Care

```{r posterior diagnostics comfort}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_comfort))

```

#### Early WLST

```{r posterior diagnostics early wlst}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_early_wlst))

```

#### DNR

```{r posterior diagnostics dnr}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_dnr_binary))

```

#### mRS 90

```{r posterior diagnostics mrs}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_mrs_90_minimal))

```

#### EuroQOL Mobility 90

```{r posterior diagnostics euro mobility}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_euro_mobility_90_canonical))

```

#### EuroQOL Self-Care 90

```{r posterior diagnostics euro selfcare}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_euro_selfcare_90_canonical))

```

#### EuroQOL Usual Activities 90

```{r posterior diagnostics euro usual}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_euro_usual_90_canonical))

```

#### EuroQOL Pain/Discomfort 90

```{r posterior diagnostics euro pain}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_euro_pain_90_canonical))

```

#### EuroQOL Anxiety/Depression 90

```{r posterior diagnostics euro anxiety}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_neutral_euro_anxiety_90_canonical))

```

#### EuroQOL VAS 90

```{r posterior diagnostics euro vas}
#| echo: false
#| fig-height: 10

with_dir(here(), tar_read(diag_posterior_vas_90))

```

### Posterior predictive checks

#### Neurosurgical Intervention

```{r posterior predictions neurosurgery}
#| echo: false

with_dir(here(), tar_read(post_pred_check_neurosurgery))

```

#### Ventriculostomy

```{r posterior predictions evd}
#| echo: false

with_dir(here(), tar_read(post_pred_check_evd))

```

#### Tracheostomy

```{r posterior predictions trach}
#| echo: false

with_dir(here(), tar_read(post_pred_check_tracheostomy))

```

#### Days of Mechanical Ventilation

```{r posterior predictions vent}
#| echo: false

with_dir(here(), tar_read(post_pred_check_days_mechanical_ventilation))

```

#### Comfort Care

```{r posterior predictions comfort}
#| echo: false

with_dir(here(), tar_read(post_pred_check_comfort))

```

#### Early WLST

```{r posterior predictions early wlst}
#| echo: false

with_dir(here(), tar_read(post_pred_check_early_wlst))

```

#### DNR

```{r posterior predictions dnr}
#| echo: false

with_dir(here(), tar_read(post_pred_check_dnr_binary))

```

#### mRS 90

```{r posterior predictions mrs}
#| echo: false

with_dir(here(), tar_read(post_pred_check_neutral_mrs_90_canonical))

```

#### EuroQOL Moblity 90

```{r posterior predictions euro mobility}
#| echo: false

with_dir(here(), tar_read(post_pred_check_neutral_euro_mobility_90_canonical))

```

#### EuroQOL Self-Care 90

```{r posterior predictions euro selfcare}
#| echo: false

with_dir(here(), tar_read(post_pred_check_neutral_euro_selfcare_90_canonical))

```

#### EuroQOL Usual Activities 90

```{r posterior predictions euro usual}
#| echo: false

with_dir(here(), tar_read(post_pred_check_neutral_euro_usual_90_canonical))

```

#### EuroQOL Pain/Discomfort 90

```{r posterior predictions euro pain}
#| echo: false

with_dir(here(), tar_read(post_pred_check_neutral_euro_pain_90_canonical))

```

#### EuroQOL Anxiety/Despression 90

```{r posterior predictions euro anxiety}
#| echo: false

with_dir(here(), tar_read(post_pred_check_neutral_euro_anxiety_90_canonical))

```

#### EuroQOL VAS 90

```{r posterior predictions euro vas}
#| echo: false

with_dir(here(), tar_read(post_pred_check_vas_90))

```

## Sensitivity Analyses

To assess the robustness of our findings for the primary outcome, we conducted several sensitivity analyses. First, we repeated the primary analysis using three alternative prior distributions: a left-hemisphere prior, a right-hemisphere prior, and a flat (minimally informative) prior. This allowed us to evaluate the influence of different initial beliefs on the resulting posterior distributions. Second, we explored an alternative covariate adjustment set based on a minimal adjustment set identified from our directed acyclic graphs (DAGs) to ensure our results were not dependent on the specific model specification. Finally, we employed an alternative method for handling missing data to confirm that our findings were not unduly influenced by the imputation strategy used in the primary analysis. The results of these sensitivity analyses were consistent with our primary findings, demonstrating the robustness of our conclusions.

### Table 2 with Different Priors

#### Left Hemisphere Prior

```{r}

with_dir(here(), tar_read(table2_left))

```

#### Right Hemisphere Prior

```{r}

with_dir(here(), tar_read(table2_right))

```

#### Flat Prior

```{r}

with_dir(here(), tar_read(table2_flat))

```

### Table 4 with Different Priors

#### Left Hemisphere Prior

```{r}

with_dir(here(), tar_read(table4_left))

```

#### Right Hemisphere Prior

```{r}

with_dir(here(), tar_read(table4_right))

```

#### Flat Prior

```{r}

with_dir(here(), tar_read(table4_flat))

```

### Alternative Handling of Missing Data

The primary analysis was conducted as a complete case analysis. We performed multiple imputation by chained equations and created 100 imputed datasets with the `mice` package[@buurenMiceMultivariateImputation2011]. Three separate Bayesian models were fitted to the 100 imputed datasets using the following outcomes: neurosurgical intervention, external ventricular drain, and modified Rankin score at 90 days. All models were fitted using Markov Chain Monte Carlo (MCMC) with 4 chains and 2,000 iterations per chain (including a warmup period of 1,000 iterations).

```{r imputed}

with_dir(here(), tar_read(table_imp))

```