---
title: "Association of Hemispheric Laterality with Neurosurgical Intervention in Intracerebral Hemorrhage: A Post-Hoc Bayesian Analysis of ATACH-2 and ERICH"
authors:
  - name: Nikhil M. Patel
    degrees: MD, MBA
    orcid: 0000-0002-3193-8590
    affiliations:
      - ref: wfsom-int-med
      - ref: wfsom-neuro
    corresponding: true
    email: nikhil.m.patel@wfusm.edu
  - name: Rahul R. Karamchandani
    degrees: MD
    affiliations:
      - ref: wfsom-neuro
  - name: Matthew N. Jaffa
    degrees: DO
    affiliations:
      - ref: hartford-neuro
  - name: Liang Wang
    degrees: PhD
    affiliations:
      - ref: advocate-analytics
  - name: Hongmei Yang
    degrees: PhD
    affiliations:
      - ref: advocate-analytics
  - name: Rhonda Cadena
    degrees: MD
    affiliations:
      - ref: wfsom-int-med
  - name: Christa Swisher
    degrees: MD
    affiliations:
      - ref: wfsom-neuro
  - name: Joe D. Bernard
    degrees: MD
    affiliations:
      - ref: cnsa
  - name: Jonathan M. Parish
    degrees: MD
    affiliations:
      - ref: cnsa
  - name: William R. Stetler
    degrees: MD
    affiliations:
      - ref: cnsa
  - name: Rick Magun
    degrees: MD
    affiliations:
      - ref: buffalo-neuro
  - name: Aaron Taylor
    degrees: MD
    affiliations:
      - ref: buffalo-neuro
  - name: Daniel Woo
    degrees: MD
    affiliations:
      - ref: buffalo-neuro
  - name: Adnan Qureshi
    degrees: MD
    affiliations:
      - ref: missouri-neuro
  - name: Andrew W. Asimos
    degrees: MD
    affiliations:
      - ref: wfsom-em
  - name: Nicholas A. Morris
    degrees: MD
    affiliations:
      - ref: umaryland-neuro

affiliations:
  - id: wfsom-int-med
    name: Department of Internal Medicine, Wake Forest School of Medicine, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: wfsom-neuro
    name: Department of Neurology, Neurosciences Institute, Wake Forest School of Medicine, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: hartford-neuro
    name: Division of Neurocritical Care, Department of Neurology, Ayer Neuroscience Institute, Hartford Hospital
    city: Hartford
    state: CT
    country: USA
  - id: advocate-analytics
    name: Clinical Quality Analytics, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: cnsa
    name: Carolina Neurosurgery & Spine Associates, Department of Neurosurgery, Neurosciences Institute, Wake Forest University School of Medicine, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: buffalo-neuro
    name: Department of Neurology, University at Buffalo Jacobs School of Medicine & Biomedical Sciences
    city: Buffalo
    state: NY
    country: USA
  - id: missouri-neuro
    name: Zeenat Qureshi Stroke Institute and Department of Neurology, University of Missouri-Columbia
    city: Columbia
    state: MO
    country: USA
  - id: wfsom-em
    name: Department of Emergency Medicine, Neurosciences Institute, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: umaryland-neuro
    name: Department of Neurology, Program in Trauma, University of Maryland School of Medicine
    city: Baltimore
    state: MD
    country: USA
format:
  docx:
    reference-doc: custom-reference.docx
bibliography: references.bib
csl: stroke.csl
table-of-contents: false
toc-title: Contents
echo: false
output: false
message: false
---

```{r setup}
library(targets)
library(tidyverse)
library(scales)
library(withr)
library(here)
library(gt)

# load data
with_dir(here(), tar_load(ich_aggressive))

## patient counts
sample_size <- count(ich_aggressive)

left_hemisphere_count <- ich_aggressive |>
  filter(ich_laterality == "Left") |>
  count()

erich_count <- ich_aggressive |>
  filter(study == "ERICH") |>
  count()

atach_count <- ich_aggressive |>
  filter(study == "ATACH-2") |>
  count()

## table 1
with_dir(here(), tar_load(table_1))

## table 2
with_dir(here(), tar_load(table_2_neutral))
tibble2 <- table_2_neutral[["_data"]]

## table 3
with_dir(here(), tar_load(table_3_neutral))
tibble3 <- table_3_neutral[["_data"]]
```

## Introduction

Intracerebral hemorrhage (ICH) remains a challenging disease, marked by its sudden onset, morbidity, and mortality. [@hemphillICHScoreSimple2001; @shethSpontaneousIntracerebralHemorrhage2022] While management of ischemic stroke has advanced considerably over the last decade, the acute management of ICH remains limited to blood pressure control, coagulopathy reversal, and more perhaps surgical evacuation of hemorrhage. [@pradillaTrialEarlyMinimally2024; @bakoMinimallyInvasiveSurgery2023] These efforts have focused on reducing mortality and preventing additional morbidity due to secondary injury.

Clinical presentations for patients with stroke are varied based on the hemisphere affected, where hemispheric dominance dictates cognitive processing, attention, language, and emotional responses. Consequently, when confronted with ICH, the specific hemisphere affected can profoundly influence not only the clinical presentation but also management decisions and prognostic outcomes.

Previous research involving patients with ischemic stroke provide evidence that those with right hemispheric injury present later to the hospital and may not be treated as frequently with thrombolytic therapy. [@dileggeImpactLesionSide2005; @foerchDifferenceRecognitionRight2005; @portegiesLeftsidedStrokesAre2015; @liStrokeLateralizationLarge2021] Some practitioners hesitate to offer decompressive craniectomy in patients with left dominant-hemispheric ischemic infarcts despite literature demonstrating that long-term functional outcomes and health-related quality of life (HRQoL) are equivalent if not better for patients with left hemispheric injuries. [@deb-chatterjiSideMattersDifferences2022; @leggeNeglectingDifference2006; @beuscherFunctionalLongTermOutcome2017; @satoHigherMortalityPatients2015]

Amidst an increasing enthusiasm for surgical management of ICH [@bakoMinimallyInvasiveSurgery2023; @pradillaTrialEarlyMinimally2024; @beckDecompressiveCraniectomyBest2024], it is important to understand whether hemispheric lateralization is associated with differences in surgical decision making, as has been observed for ischemic stroke. We aimed to evaluate the association of hemispheric lateralization of ICH with the likelihood of receiving neurosurgical intervention as well as with functional outcomes.

## Methods

### Study Design and Participants

We conducted a post-hoc analysis using data from two prospective studies: the Antihypertensive Treatment of Acute Cerebral Hemorrhage II (ATACH-2) trial and the Ethnic/Racial Variations of Intracerebral Hemorrhage (ERICH) study. ATACH-2 was a large, multicenter, international, open-label, blinded-endpoint randomized controlled trial that enrolled 1000 patients with ICH and randomized them to either intensive blood pressure control (110-139 mm Hg) or standard blood pressure control (140-179 mm Hg). ERICH was a large, multicenter, prospective observational study conducted patients in the United States on patients admitted with ICH, with particular emphasis on recruitment of a racially and ethnically diverse cohort. The protocols and results of both studies have been previously published elsewhere [@qureshiIntensiveBloodPressureLowering2016; @wooEthnicRacialVariations2013].

We included all patients enrolled in these two studies with supratentorial location of ICH. We excluded patients with cerebellar or brainstem ICH as well as any patient with primary intraventricular hemorrhage (IVH).

The study was approved by the institutional review board of the Wake Forest School of Medicine (IRB00098198).

### Exposures and Outcomes

Our exposure was hemispheric laterality (left versus right). Our primary outcome was neurosurgical intervention as a binary variable defined as receiving a hematoma evacuation, decompressive craniectomy, or other craniotomy.

To explore differences in other aspects of aggressive care, we defined several secondary outcomes: receipt of external ventricular drain (EVD), receipt of tracheostomy, days of mechanical ventilation, receipt of do-not-resuscitate (DNR) order at any point in the hospitalization, withdrawal of life-sustaining therapy (WLST) at any point in the hospitalization, and early WLST (defined as \< 72 from symptom onset).

Our primary functional outcome was modified Rankin score (mRS) at 90 days analyzed via ordinal regression. We also examined quality of life as assessed by the European Quality of Life-5 dimensions (EQ-5D) questionnaire.

### Statistical Analysis

To estimate the effect of hemispheric laterality on the outcomes of interest, we fitted hierarchal Bayesian regression models adjusted for the specified covariates below. We also included the study that patient came from as a random intercept. We employed Bayesian statistical methods for this analysis due to several advantages over traditional frequentist approaches. Bayesian methods allow for the incorporation of prior knowledge or beliefs into the analysis, which can be particularly useful in observational studies where prior information about relationships between variables may exist. This is achieved through the use of prior distributions, which are combined with the observed data to produce posterior distributions representing updated beliefs about the parameters of interest. Furthermore, Bayesian methods provide a natural framework for quantifying uncertainty through credible intervals (CrI), which offer a more intuitive interpretation than frequentist confidence intervals. Finally, Bayesian modeling facilitates the estimation of posterior probabilities, allowing for direct probabilistic statements about hypotheses, such as the probability that an effect size (e.g., odds ratio [OR] or incidence rate ratio [IRR]) is greater than 1, which is often of greater clinical relevance than a p-value. [@dogucuBayesRulesIntroduction; @mcelreathStatisticalRethinkingBayesian2020] This report adheres to the Bayesian Analysis Reporting Guidelines. [@kruschkeBayesianAnalysisReporting2021] 

For binary and ordinal endpoints, we utilized hierarchical Bayesian logistic and ordinal logistic regression models, respectively. For non-binary outcomes, we employed specialized distributional models to avoid data dichotomization and accurately reflect the underlying data generating processes. Specifically, days of mechanical ventilation, a count-based outcome characterized by excess zeros, was modeled using a hierarchical Bayesian zero-inflated negative binomial regression. Exponentiated coefficients for the count component of this model are reported as incidence rate ratios (IRR). The European Quality of Life Visual Analog Scale (EQ-VAS), a bounded continuous measure, was scaled to a proportion and modeled using a hierarchical Bayesian zero-one inflated beta (ZOIB) regression. To maximize clinical interpretability, we applied G-computation to derive the average marginal expected mean difference on the original 0–100 point scale, rather than reporting proportional odds. 

Decisions on neurosurgical intervention and other aspects of acute care are highly influenced by local practice patterns and resources. To account for potential clustering of clinical decision-making by hospital, we ideally would have included a random intercept for the study site in all models. However, site-level identifiers were not available for the ERICH cohort. To assess the potential impact of omitting site-level clustering, we performed a sensitivity analysis restricted to the ATACH-2 cohort, where site identifiers were available. We fitted the primary model for neurosurgical intervention with and without a random intercept for the hospital site and compared the posterior estimates for the association with laterality.

All analyses were performed in R version 4.4.2 (R Foundation for Statistical Computing, Vienna, Austria). ALl packages used for the analysis are listed in the Supplementary Appendix.

#### Covariate Selection

We constructed directed acyclic graphs (DAGs) to model our causal assumptions and to guide covariate selection for each estimand. [@debertinSynthesizingSubjectmatterExpertise2024; @tennantUseDirectedAcyclic2021; @hernanSpecifyingTargetTrial2016] For neurosurgical intervention and other acute care outcomes, the target estimand was the direct effect of hemispheric laterality on treatment decisions. Models were adjusted for baseline confounders that independently influence both the clinical presentation associated with hemispheric laterality and the likelihood of intervention: age, admission Glasgow Coma Scale (GCS) score, ICH location (categorized as lobar, thalamic, or basal ganglia), baseline ICH volume, and presence of intraventricular hemorrhage (IVH).

For functional outcomes (mRS and EQ-5D), the target estimand was the total effect of hemispheric laterality, encompassing all causal pathways including those mediated through neurosurgical intervention and other post-baseline care decisions. These models were adjusted for the same baseline confounders but did not condition on neurosurgical intervention, as it lies on the causal pathway from laterality to functional outcome; adjusting for a mediator would block the clinically relevant pathway and risk introducing collider stratification bias. [@tennantUseDirectedAcyclic2021] The DAGs and further details of the covariate selection rationale are presented in the supplementary material.

#### Prior Distribution Specification

We applied four sets of prior distributions for all models to assess the robustness of our findings. The first set was a neutral prior, centered on a log-odds ratio of 0, which corresponds to an odds ratio of 1 and assumes no difference in the outcome of interest between left and right hemispheric patients. The second and third sets were left hemisphere and right hemisphere priors, respectively, which were weakly informative priors that assumed a higher probability of the outcome occurring in the specified hemisphere. The final set was a flat prior, which assumes that all possible values for the model parameters are equally likely. This non-informative prior allows the data to dominate the posterior distribution and most closely approximates the assumptions of a frequentist analysis. More details on the rationale of prior specification and the specific parameters used for each prior distribution are included in the supplementary appendix.

#### Computation and Description of the Posterior Distribution

We report the following metrics for the effect of hemispheric laterality: 1) median posterior effect size (OR, IRR, or mean difference) with 95% CrI; 2) probability of any directional difference (e.g., P\[OR\] \> 1 or P\[Mean Difference\] \< 0); 3) probability of a substantial difference defined *a priori* (e.g., P\[OR\] \> 1.2 or P\[Mean Difference\] \< -5 points); and 4) percentage of the posterior distribution falling outside a predefined region of practical equivalence (ROPE; e.g., 0.95 \< OR \< 1.05, or ±2 points for mean differences). We defined a substantial difference as an OR \> 1.2, which represents a small but clinically meaningful difference in effect size commonly utilized in epidemiological research. The full rationale for these thresholds is detailed in the supplementary material.

#### Missing Data Treatment and Sensitivity Analyses

Baseline covariates used in model adjustment had minimal missingness. Missingness for 90-day outcomes was more substantial but was balanced across laterality groups and key covariates. Detailed missingness rates by laterality group, assessments supporting the Missing at Random (MAR) assumption, and imputation diagnostics are reported in the Supplementary Appendix.

Our primary analysis used multiple imputation by chained equations (MICE) to address missing outcome data, generating 20 imputed datasets with the `mice` package.[@buurenMiceMultivariateImputation2011] Imputation models included all analysis covariates and outcomes to preserve inter-variable relationships under the MAR assumption. Predictive mean matching was used for continuous and ordinal variables, logistic regression for binary variables, and polytomous regression for nominal categorical variables. Each Bayesian model was fitted separately to each imputed dataset, and posterior draws were pooled across imputations to produce the reported estimates and CrI.

In addition to the missing data analysis, we analyzed our models using the four sets of prior distributions described above, which allowed us to examine the influence of prior choices on the posterior distributions. To explore potential effect modification, we also fitted models that included formal interaction terms between hemispheric laterality and ICH location and between laterality and study source. Subgroup-specific aORs and interaction Ratio of Odds Ratios (ROR) with 95% CrI were computed from these models to evaluate whether the laterality effect differed across subgroups. Further details on the methods and results of these sensitivity and subgroup analyses are presented in the Supplementary Appendix.

### Data Availability Statement

The ATACH-2 (PI: Adnan I. Qureshi, MD, NCT01176565) and ERICH (PI: Daniel Woo, MD, MS, NCT01202864) data and analyses presented in this manuscript are based on the research files downloaded from the Archived Clinical Research web site of the National Institute of Neurologic Disease and Stroke (NINDS). The datasets can be obtained after approval of a written proposal as outlined on the NINDS website. As outlined in the NINDS Data Request Agreement, the analysis code is publicly available at <https://github.com/nikhilpatelmd/ich-laterality>.

## Results

### Baseline Characteristics

A total of `r sample_size %>% pull(n)` patients were included in the analysis, of which `r left_hemisphere_count %>% pull(n)` (`r percent(left_hemisphere_count %>% pull(n) / sample_size %>% pull(n), digits = 1)`) had left hemispheric ICH (Table 1). They had higher National Institutes of Health Stroke Scale (NIHSS) scores, lower GCS scores, and lower rates of hypertension as compared to those with right hemispheric ICH. There were `r erich_count %>% pull(n)` (`r percent(erich_count %>% pull(n) / sample_size %>% pull(n), digits = 1)`) patients in the cohort from ERICH and `r atach_count %>% pull(n)` (`r percent(atach_count %>% pull(n) / sample_size %>% pull(n), digits = 1)`) from ATACH-2.

### Hemispheric Laterality and Neurosurgical Intervention

```{r neurosurgical-intervention}
# Extract the values for the neurosurgical intervention outcome
or_ci_values_surg <- tibble2 %>%
  filter(Outcome == "Neurosurgical Intervention") %>%
  pull(or_ci)

or_ci_split_surg <- str_split(or_ci_values_surg, " \\(| - ", simplify = TRUE)

odds_ratio_surgery <- or_ci_split_surg[[1]]
lower_bound_surg <- or_ci_split_surg[[2]]
upper_bound_surg <- str_remove(or_ci_split_surg[[3]], "\\)")
```

Associations between hemispheric laterality and primary and secondary outcomes are shown in Table 2. `r tibble2 %>% filter(Outcome == "Neurosurgical Intervention") %>% pull(Left)` patients with left hemispheric ICH underwent neurosurgical intervention, compared to `r tibble2 %>% filter(Outcome == "Neurosurgical Intervention") %>% pull(Right)` with right hemispheric ICH (aOR `r odds_ratio_surgery`, 95% CrI `r lower_bound_surg`–`r upper_bound_surg`). The posterior probability that the aOR exceeded 1 was `r tibble2 %>% filter(Outcome == "Neurosurgical Intervention") %>% pull(or_1)`, and the posterior probability of a substantial difference (aOR > 1.2) was `r tibble2 %>% filter(Outcome == "Neurosurgical Intervention") %>% pull(or_1.2)`. Fewer than `r tibble2 %>% filter(Outcome == "Neurosurgical Intervention") %>% pull(rope)` of the posterior distribution fell within the ROPE.

Figure 1 displays the covariate-adjusted marginal predicted probabilities of neurosurgical intervention and the prior sensitivity analysis. Panel A shows the posterior distributions of covariate-adjusted absolute predicted probabilities of neurosurgical intervention for left and right hemispheric ICH under the neutral prior (left hemisphere: 4.86%, 95% CrI 3.98–5.84%; right hemisphere: 8.04%, 95% CrI 6.93–9.23%). Panel B shows the posterior distributions of the absolute probability difference (Right − Left) across all four prior specifications. Under the neutral prior, the median probability difference was 3.18 percentage points (95% CrI 1.75–4.62), with P(Right > Left) exceeding 99.9%. Corresponding estimates were 2.21 percentage points (95% CrI 0.97–3.45) under the left-hemisphere prior, 2.84 percentage points (95% CrI 1.60–4.09) under the right-hemisphere prior, and 3.38 percentage points (95% CrI 1.90–4.89) under the flat prior, with P(Right > Left) exceeding 99.9% under all specifications.

In a sensitivity analysis restricted to the ATACH-2 cohort, inclusion of a random intercept for hospital site modestly attenuated the point estimate and shifted the credible interval toward the null but did not materially alter the direction or strength of the association. Details are available in the Supplementary Appendix.

### Hemispheric Laterality and Secondary Outcomes

```{r secondary-outcomes}
# EVD
or_ci_values_evd <- tibble2 %>%
  filter(Outcome == "External Ventricular Drain") %>%
  pull(or_ci)

or_ci_split_evd <- str_split(or_ci_values_evd, " \\(| - ", simplify = TRUE)

odds_ratio_evd <- or_ci_split_evd[[1]]
lower_bound_evd <- or_ci_split_evd[[2]]
upper_bound_evd <- str_remove(or_ci_split_evd[[3]], "\\)")

# Mechanical ventilation (count outcome; reported as IRR)
irr_ci_values_mv <- tibble2 %>%
  filter(Outcome == "Days of Mechanical Ventilation") %>%
  pull(or_ci)

irr_ci_split_mv <- str_split(irr_ci_values_mv, " \\(| - ", simplify = TRUE)

irr_mv <- irr_ci_split_mv[[1]]
lower_bound_mv <- irr_ci_split_mv[[2]]
upper_bound_mv <- str_remove(irr_ci_split_mv[[3]], "\\)")
```

A total of `r tibble2 %>% filter(Outcome == "External Ventricular Drain") %>% pull(Left)` with left hemispheric ICH had an EVD placed as compared to `r tibble2 %>% filter(Outcome == "External Ventricular Drain") %>% pull(Right)` with right hemispheric ICH (aOR `r odds_ratio_evd`, 95% CrI `r lower_bound_evd`–`r upper_bound_evd`). The posterior probability that the aOR exceeded 
1 was `r tibble2 %>% filter(Outcome == "External Ventricular Drain") %>% pull(or_1)`, and the posterior probability of a substantial difference (aOR > 1.2) was `r tibble2 %>% filter(Outcome == "External Ventricular Drain") %>% pull(or_1.2)`. `r tibble2 %>% filter(Outcome == "External Ventricular Drain") %>% pull(rope)` of the posterior distribution fell within the ROPE.

Median days on mechanical ventilation were `r tibble2 %>% filter(Outcome == "Days of Mechanical Ventilation") %>% pull(Left)` and `r tibble2 %>% filter(Outcome == "Days of Mechanical Ventilation") %>% pull(Right)` for left and right hemispheric ICH, respectively (IRR `r irr_mv`, 95% CrI `r lower_bound_mv`–`r upper_bound_mv`). The posterior probability that the IRR exceeded 1 was `r tibble2 %>% filter(Outcome == "Days of Mechanical Ventilation") %>% pull(or_1)`, and `r tibble2 %>% filter(Outcome == "Days of Mechanical Ventilation") %>% pull(rope)` of the posterior distribution fell within the ROPE.

Rates of tracheostomy were similar between groups (`r tibble2 %>% filter(Outcome == "Tracheostomy") %>% pull(Left)` left vs. `r tibble2 %>% filter(Outcome == "Tracheostomy") %>% pull(Right)` right; aOR `r tibble2 %>% filter(Outcome == "Tracheostomy") %>% pull(or_ci)`), with a posterior probability that the aOR exceeded 1 of `r tibble2 %>% filter(Outcome == "Tracheostomy") %>% pull(or_1)` and `r tibble2 %>% filter(Outcome == "Tracheostomy") %>% pull(rope)` of the posterior distribution within the ROPE. Rates of WLST (`r tibble2 %>% filter(Outcome == "Withdrawal of Life-Sustaining Therapy") %>% pull(Left)` left vs. `r tibble2 %>% filter(Outcome == "Withdrawal of Life-Sustaining Therapy") %>% pull(Right)` right; aOR `r tibble2 %>% filter(Outcome == "Withdrawal of Life-Sustaining Therapy") %>% pull(or_ci)`), early WLST (`r tibble2 %>% filter(Outcome == "Early Withdrawal of Life-Sustaining Therapy") %>% pull(Left)` left vs. `r tibble2 %>% filter(Outcome == "Early Withdrawal of Life-Sustaining Therapy") %>% pull(Right)` right; aOR `r tibble2 %>% filter(Outcome == "Early Withdrawal of Life-Sustaining Therapy") %>% pull(or_ci)`), and DNR orders (`r tibble2 %>% filter(Outcome == "DNR Order") %>% pull(Left)` left vs. `r tibble2 %>% filter(Outcome == "DNR Order") %>% pull(Right)` right; aOR `r tibble2 %>% filter(Outcome == "DNR Order") %>% pull(or_ci)`) were also similar across hemispheres, with P(aOR > 1) of `r tibble2 %>% filter(Outcome == "Withdrawal of Life-Sustaining Therapy") %>% pull(or_1)`, `r tibble2 %>% filter(Outcome == "Early Withdrawal of Life-Sustaining Therapy") %>% pull(or_1)`, and `r tibble2 %>% filter(Outcome == "DNR Order") %>% pull(or_1)`, respectively.

### Hemispheric Laterality and Functional Outcomes

```{r functional-outcomes}
# mRS
or_ci_values_mrs <- tibble3 %>%
  filter(outcome == "Modified Rankin Score") %>%
  pull(est_label)

or_ci_split_mrs <- str_split(or_ci_values_mrs, " \\(| - ", simplify = TRUE)

odds_ratio_mrs <- or_ci_split_mrs[[1]]
lower_bound_mrs <- or_ci_split_mrs[[2]]
upper_bound_mrs <- str_remove(or_ci_split_mrs[[3]], "\\)")

# EuroQOL - Mobility
or_ci_values_mob <- tibble3 %>%
  filter(outcome == "EuroQOL - Mobility") %>%
  pull(est_label)

or_ci_split_mob <- str_split(or_ci_values_mob, " \\(| - ", simplify = TRUE)

odds_ratio_mob <- or_ci_split_mob[[1]]
lower_bound_mob <- or_ci_split_mob[[2]]
upper_bound_mob <- str_remove(or_ci_split_mob[[3]], "\\)")

# Euro VAS
or_ci_values_vas <- tibble3 %>%
  filter(outcome == "Euro VAS") %>%
  pull(est_label)

or_ci_split_vas <- str_split(or_ci_values_vas, " \\(| - ", simplify = TRUE)

mean_diff_vas <- or_ci_split_vas[[1]]
lower_bound_vas <- or_ci_split_vas[[2]]
upper_bound_vas <- str_remove(or_ci_split_vas[[3]], "\\)")
```

Associations between hemispheric laterality and functional outcomes are shown in Table 3. The aOR for a worse mRS score at 90 days for right hemispheric ICH compared to left hemispheric ICH was `r odds_ratio_mrs` (95% CrI `r lower_bound_mrs`–`r upper_bound_mrs`), with a posterior probability that the aOR exceeded 1 of `r tibble3 %>% filter(outcome == "Modified Rankin Score") %>% pull(prob_diff)` and `r tibble3 %>% filter(outcome == "Modified Rankin Score") %>% pull(prob_rope)` of the posterior distribution within the ROPE. Figure 3 displays the covariate-adjusted absolute marginal predicted cumulative distributions of mRS at 90 days for left and right hemispheric ICH under the neutral prior.

For Mobility and Self-Care, the posterior probability that the aOR exceeded 1 was `r tibble3 %>% filter(outcome == "EuroQOL - Mobility") %>% pull(prob_diff)` for both domains (aOR `r odds_ratio_mob`, 95% CrI `r lower_bound_mob`–`r upper_bound_mob` for Mobility; aOR `r tibble3 %>% filter(outcome == "EuroQOL - Self-Care") %>% pull(est_label)` for Self-Care), with a posterior probability of a substantial difference (aOR > 1.2) of `r tibble3 %>% filter(outcome == "EuroQOL - Mobility") %>% pull(prob_sub)` for both, and `r tibble3 %>% filter(outcome == "EuroQOL - Mobility") %>% pull(prob_rope)` and `r tibble3 %>% filter(outcome == "EuroQOL - Self-Care") %>% pull(prob_rope)` of the posterior distributions within the ROPE, respectively. For Pain/Discomfort, the aOR was `r tibble3 %>% filter(outcome == "EuroQOL - Pain/Discomfort") %>% pull(est_label)` with P(aOR > 1) of `r tibble3 %>% filter(outcome == "EuroQOL - Pain/Discomfort") %>% pull(prob_diff)` and `r tibble3 %>% filter(outcome == "EuroQOL - Pain/Discomfort") %>% pull(prob_rope)` of the posterior distribution within the ROPE. For Anxiety/Depression, the aOR was `r tibble3 %>% filter(outcome == "EuroQOL - Anxiety/Depression") %>% pull(est_label)` with P(aOR > 1) of `r tibble3 %>% filter(outcome == "EuroQOL - Anxiety/Depression") %>% pull(prob_diff)` and `r tibble3 %>% filter(outcome == "EuroQOL - Anxiety/Depression") %>% pull(prob_rope)` of the posterior distribution within the ROPE. For Usual Activities, the aOR was `r tibble3 %>% filter(outcome == "EuroQOL - Usual Activities") %>% pull(est_label)` with P(aOR > 1) of `r tibble3 %>% filter(outcome == "EuroQOL - Usual Activities") %>% pull(prob_diff)` and `r tibble3 %>% filter(outcome == "EuroQOL - Usual Activities") %>% pull(prob_rope)` of the posterior distribution within the ROPE. For the Euro VAS, the estimated mean difference was `r mean_diff_vas` points (95% CrI `r lower_bound_vas`–`r upper_bound_vas`), with a posterior probability of `r tibble3 %>% filter(outcome == "Euro VAS") %>% pull(prob_diff)` that right hemispheric ICH was associated with a lower score, `r tibble3 %>% filter(outcome == "Euro VAS") %>% pull(prob_rope)` of the posterior distribution within the ROPE, and a posterior probability of a substantial difference (mean difference < −5 points) of `r tibble3 %>% filter(outcome == "Euro VAS") %>% pull(prob_sub)`. Results from sensitivity analyses using alternative priors are presented in the Supplementary Appendix and are consistent with these results.

## Discussion

In this pooled analysis of `r sample_size` patients with ICH, we observed an association between right hemispheric ICH and a higher likelihood of receiving neurosurgical intervention compared to left hemispheric ICH. Specifically, patients with right hemispheric ICH had `r odds_ratio_surgery` times the adjusted odds of receiving a neurosurgical intervention and `r odds_ratio_evd` times the adjusted odds of receiving an EVD. This finding was robust to prior specification: all four prior distributions yielded nearly identical posteriors with P(Right > Left) exceeding 99.9%, and only marginal shifts in the posterior median (Figure 1). In exploratory subgroup analyses with formal interaction terms, the laterality association was directionally consistent across ICH locations and between the two source studies; however, the interaction Ratio of Odds Ratios (ROR) credible intervals were wide and included 1, precluding definitive conclusions about subgroup heterogeneity (Supplementary Appendix). Despite a lower observed rate of surgical intervention, patients with left hemispheric ICH had similar 90-day functional outcomes to those with right hemispheric ICH, a finding consistent across definitions including the mRS and HRQoL domains. Of note, our functional outcome models estimated the total association of hemispheric laterality with outcome — inclusive of downstream differences in surgical management — rather than the direct association conditional on treatment received, thereby reflecting the overall prognostic profile of laterality as encountered in clinical practice.

This finding is noteworthy in light of recent evidence suggesting that surgical interventions, such as hematoma evacuation and decompressive craniectomy, may improve outcomes in some ICH patients [@pradillaTrialEarlyMinimally2024; @beckDecompressiveCraniectomyBest2024]. In both the ATACH-2 and ERICH studies, neurosurgical candidacy was not defined by strict trial protocols but was left to the discretion of the treating clinicians based on local institutional practices. Decisions regarding hematoma evacuation or decompressive craniectomy were inherently influenced by temporal trends and site-specific protocols, which varied considerably over the study periods. While we adjusted for key baseline covariates and included a study-level random effect, these institutional and temporal practice variations likely shaped the overall landscape of surgical care observed in our cohort.

The observed association between laterality and neurosurgical intervention may reflect several mechanisms, which our observational design cannot fully disentangle. One possibility is a treatment selection bias in which clinicians, patients, or families preferentially avoid surgical intervention for dominant-hemispheric injuries due to concerns about post-operative language deficits. Indeed, several studies have found that clinicians, patients, and families place particular importance on language function and are less optimistic about dominant-hemispheric injuries. [@neugebauerDESTINYSAttitudesPhysicians2014; @neugebauerAttitudesPatientsRelatives2017; @neugebauerAttitudesNursesDisability2019] Prognostic framing that emphasizes aphasia may also affect surrogate decision-making and goals-of-care discussions, which in turn may influence willingness to pursue surgical intervention.

However, alternative explanations should also be considered. Residual confounding from unmeasured clinical factors — such as proximity to eloquent cortex, degree of mass effect, extent of hydrocephalus, midline shift, or more granular anatomical features beyond our lobar-versus-deep classification — may independently influence both perceived surgical candidacy and laterality-associated outcomes. Similarly, family preferences, cultural attitudes toward disability, and language-related prognostic framing could influence treatment decisions in ways that are associated with hemispheric laterality but are not captured in our dataset. Our models adjusted for key measured confounders including ICH volume, location, GCS, and IVH, but the observational and post-hoc nature of this analysis means that residual confounding cannot be excluded. Accordingly, the association we observed should not be interpreted as definitive evidence of treatment inequity, but rather as a signal warranting prospective investigation. If confirmed, this disparity could have meaningful clinical implications — patients with dominant-hemispheric ICH may report equivalent quality of life outcomes compared to those with non-dominant hemispheric injury, and prevailing assumptions about poor prognosis in this group may not be warranted. Furthermore, the observation that functional outcomes were similar between groups despite less aggressive care for left hemispheric ICH is notable. This equivalence could be partly due to the differential impact of specific neurocognitive deficits on standardized rating scales; for example, severe neglect associated with right hemispheric lesions can profoundly impair mobility and self-care domains, whereas aphasia from left hemispheric lesions may influence mRS and caregiver perceptions of disability in different ways.

Evidence is limited regarding the association of hemispheric lateralization and functional outcomes in patients with ICH. Our findings corroborate those from a pooled analysis of the INTERACT1 and INTERACT2 trials, which showed a higher risk of mortality in patients with right hemispheric ICH. [@satoHigherMortalityPatients2015] Most other studies examining the association of hemispheric laterality and functional outcomes have focused on patients with acute ischemic stroke with mixed results. The two largest studies were pooled analyses from prospective ischemic stroke trials and neither found a clear association between hemispheric laterality and outcomes. [@finkDoesHemisphericLateralization2008; @almekhlafiStrokeLateralityDid2019] This may be particularly challenging to study in ischemic stroke due to how the NIHSS is used to guide treatment decisions and is also biased toward higher scores in patients with left hemispheric injury. [@finkAssociationNationalInstitutes2002]

Our study has some limitations. As a post-hoc observational analysis, our findings demonstrate associations and cannot establish causality. We used left versus right hemispheric laterality as a proxy for hemispheric dominance; however, approximately 5–10% of the general population — and a higher proportion of left-handed individuals — have atypical (right or bilateral) language dominance, and handedness data were not available in either cohort. As a result, some degree of misclassification of dominant versus non-dominant hemispheric injury is expected, which would bias effect estimates toward the null. Residual confounding from unmeasured clinical variables — including more granular measures of hematoma anatomy, eloquent cortex involvement, mass effect, and hydrocephalus — may account for some or all of the observed differences. Our primary models included a random intercept for study (ATACH-2 versus ERICH) but could not account for site-level clustering across all centers because site identifiers were unavailable for the ERICH cohort; a sensitivity analysis restricted to ATACH-2 incorporating a site-level random intercept yielded qualitatively similar results, but residual confounding from unmeasured center-level practice variation in the full cohort cannot be excluded. The absolute number of patients undergoing surgery was relatively low, making it unclear how these findings might translate to settings with higher intervention rates. We derived our cohort from two prospective studies where practice patterns may have been affected by the Hawthorne effect. Our cohort is also limited to mostly patients living in the United States, and there is likely to be a range of different socioeconomic and cultural frameworks that factor into decisions on treatment, prognosis, and quality of life throughout the world. Additionally, we could not assess unmeasured factors such as family preferences, language-related prognostic framing by clinicians, or the specific content of goals-of-care discussions, all of which may influence treatment decisions in laterality-dependent ways. With the recent publication of several large RCTs demonstrating the benefit of hematoma evacuation and decompressive craniectomy in ICH [@pradillaTrialEarlyMinimally2024; @beckDecompressiveCraniectomyBest2024], it will be imperative to confirm our findings with additional cohorts. Future research should focus on prospectively evaluating the association between hemispheric laterality and treatment decisions, and on surveying clinicians, patients, and families with a variety of cultural and socioeconomic backgrounds to better understand the factors that drive observed differences in intervention rates.

## Conclusions

In this post-hoc observational analysis, left hemispheric ICH was associated with a lower likelihood of neurosurgical intervention and external ventricular drain placement compared to right hemispheric ICH. Despite this difference in aggressive care, left hemispheric ICH was not associated with worse functional outcomes at 90 days, with patients exhibiting similar mRS and HRQoL profiles. While these associations are consistent with possible laterality-related differences in treatment decisions, residual confounding from unmeasured clinical and social factors cannot be excluded. These findings highlight the need for prospective studies to clarify whether hemispheric laterality independently influences treatment allocation and to ensure equitable access to surgical intervention in ICH.

{{< pagebreak >}}

## References

::: {#refs}
:::

{{< pagebreak >}}

## Tables

```{r}
#| echo: false
#| output: true
#| tbl-cap: "Demographic and Clinical Characteristics of Patients with Left and Right Hemispheric Intracerebral Hemorrhage at Baseline"
#| label: tbl-1
table_1
```

{{< pagebreak >}}


```{r}
#| echo: false
#| output: true
#| tbl-cap: "Association of Hemispheric Laterality with Primary and Secondary Outcomes"
#| label: tbl-2
table_2_neutral
```

{{< pagebreak >}}


```{r}
#| echo: false
#| output: true
#| tbl-cap: "Association of Hemispheric Laterality with Functional Outcomes"
#| label: tbl-3
table_3_neutral
```

{{< pagebreak >}}

## Figures

![Prior Sensitivity Analysis for Neurosurgical Intervention](../figures/figure_sensitivity_neurosurgery_evac.pdf){#fig-1}

![modified Rankin Score at 90 days Stratified by ICH Laterality](../figures/mrs/figure_mrs_main_neutral_adjusted.pdf){#fig-2}
