---
title: "Association of Hemispheric Laterality with Neurosurgery in ICH: A Post-Hoc Bayesian Analysis of ATACH-2 and ERICH"
authors:
  - name: Nikhil M. Patel
    degrees: MD, MBA
    orcid: 0000-0002-3193-8590
    affiliations:
      - ref: wfsom-int-med
      - ref: wfsom-neuro
    corresponding: true
    email: nikhil.m.patel@wfusm.edu
  - name: Rahul R. Karamchandani
    degrees: MD
    affiliations:
      - ref: wfsom-neuro
  - name: Matthew N. Jaffa
    degrees: DO
    affiliations:
      - ref: hartford-neuro
  - name: Liang Wang
    degrees: PhD
    affiliations:
      - ref: advocate-analytics
  - name: Hongmei Yang
    degrees: PhD
    affiliations:
      - ref: advocate-analytics
  - name: Rhonda Cadena
    degrees: MD
    affiliations:
      - ref: wfsom-int-med
  - name: Christa Swisher
    degrees: MD
    affiliations:
      - ref: wfsom-neuro
  - name: Joe D. Bernard
    degrees: MD
    affiliations:
      - ref: cnsa
  - name: Jonathan M. Parish
    degrees: MD
    affiliations:
      - ref: cnsa
  - name: William R. Stetler
    degrees: MD
    affiliations:
      - ref: cnsa
  - name: Rick Magun
    degrees: MD
    affiliations:
      - ref: buffalo-neuro
  - name: Aaron Taylor
    degrees: MD
    affiliations:
      - ref: buffalo-neuro
  - name: Daniel Woo
    degrees: MD
    affiliations:
      - ref: buffalo-neuro
  - name: Adnan Qureshi
    degrees: MD
    affiliations:
      - ref: missouri-neuro
  - name: Andrew W. Asimos
    degrees: MD
    affiliations:
      - ref: wfsom-em
  - name: Nicholas A. Morris
    degrees: MD
    affiliations:
      - ref: umaryland-neuro

affiliations:
  - id: wfsom-int-med
    name: Department of Internal Medicine, Wake Forest School of Medicine, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: wfsom-neuro
    name: Department of Neurology, Neurosciences Institute, Wake Forest School of Medicine, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: hartford-neuro
    name: Division of Neurocritical Care, Department of Neurology, Ayer Neuroscience Institute, Hartford Hospital
    city: Hartford
    state: CT
    country: USA
  - id: advocate-analytics
    name: Clinical Quality Analytics, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: cnsa
    name: Carolina Neurosurgery & Spine Associates, Department of Neurosurgery, Neurosciences Institute, Wake Forest University School of Medicine, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: buffalo-neuro
    name: Department of Neurology, University at Buffalo Jacobs School of Medicine & Biomedical Sciences
    city: Buffalo
    state: NY
    country: USA
  - id: missouri-neuro
    name: Zeenat Qureshi Stroke Institute and Department of Neurology, University of Missouri-Columbia
    city: Columbia
    state: MO
    country: USA
  - id: wfsom-em
    name: Department of Emergency Medicine, Neurosciences Institute, Advocate Health
    city: Charlotte
    state: NC
    country: USA
  - id: umaryland-neuro
    name: Department of Neurology, Program in Trauma, University of Maryland School of Medicine
    city: Baltimore
    state: MD
    country: USA
format:
  docx:
    reference-doc: custom-reference.docx
bibliography: references.bib
csl: stroke.csl

---

```{r setup}
#| echo: false
#| output: false
#| fig-height: 12
library(targets)
library(tidyverse)
library(withr)
library(here)
library(ggdag)
library(dagitty)
library(modelsummary)
library(marginaleffects)
library(brms)
library(tidybayes)
library(bayesplot)
library(glue)
library(ggdist)
library(patchwork)
library(tinytable)
library(gt)


# plot stuff
theme_ich <- function(base_family = "Figtree", base_size = 10) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(lineheight = 2),
      plot.title = element_text(size = rel(1.3), margin = margin(12, 0, 8, 0)),
      plot.subtitle = element_text(
        size = rel(1.1),
        margin = margin(4, 0, 0, 0)
      ),
      axis.text.y = element_text(size = rel(0.8)),
      axis.title.y = element_text(size = 12, margin = margin(0, 4, 0, 0)),
      axis.text.x = element_text(size = rel(2)),
      axis.title.x = element_blank(),
      legend.position = "bottom",
      legend.justification = 1,
      panel.grid = element_line(colour = "#F3F4F5"),
      plot.caption = element_text(size = rel(0.8), margin = margin(8, 0, 0, 0)),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25, "cm")
    )
}

# load data and models
with_dir(here(), tar_load(ich_aggressive))
```

# Details of Statistical Analysis

## Covariate Selection

As mentioned in the primary manuscript, we constructed directed acyclic graphs (DAGs) to facilitate covariate selection. DAGs are graphical models used to represent causal relationships between variables. They consist of nodes (representing variables) and directed arrows (representing causal influences). DAGs can help minimize several sources of bias that are inherent in many observational studies, such as confounding bias and selection bias. When constructed with subject-matter knowledge, they can provide a framework for making causal inferences from observational data. Specifically, they can help researchers with three key benefits:

1. **Identify minimal adjustment sets**: Determine the smallest set of variables to control for to estimate the causal effect of an exposure on an outcome. 

2. **Avoid over-adjustment**: Prevent controlling for variables that are mediators or colliders, which can bias the results.

3. **Clarify causal assumptions**: Make the underlying causal assumptions explicit and transparent. [@hernanSpecifyingTargetTrial2016]

In constructing DAGs for this project, we used guidance and recommendations from previous literature. [@tennantUseDirectedAcyclic2021; @sauerUseDirectedAcyclic2013; @barrettCausalInference2025]

### Primary and Secondary Outcomes

For our primary outcome model we specified the following DAG:

```{r neurosurgery dag}
#| echo: false
#| warning: false
#| fig-height: 6
library(ggokabeito)

with_dir(here(), tar_load(dag_neurosurgery))

status_colors <- c(exposure = "#6f86d6", outcome = "#764ba2", latent = "grey50")

dag_neurosurgery |>
  tidy_dagitty() |>
  node_status() |>
  ggdag() +
  geom_dag_edges() +
  geom_dag_point(aes(color = status)) +
  geom_dag_label_repel(
    aes(label = label, fill = status),
    seed = 1234,
    color = "white",
    fontface = "bold"
  ) +
  scale_color_manual(values = status_colors, na.value = "grey20") +
  scale_fill_manual(values = status_colors, na.value = "grey20") +
  guides(color = "none", fill = "none") +
  theme_dag()
```

Before considering if covariate adjustment is indicated for our outcome of interest, we need to establish the difference between the total, direct, and indirect effect of hemispheric laterality on neurosurgical intervention and the effect with which we are interested. [@tennantUseDirectedAcyclic2021]

The **total effect** of hemispheric laterality (X) on neurosurgical intervention (Y) refers to the overall influence of X on Y, regardless of the mechanisms through which that influence operates. It is the effect we would observe if we simply compared outcomes between patients with left vs. right hemispheric ICH without controlling for any variables.

The **direct effect** of X on Y refers to the influence of X on Y that is not mediated by any other variables in the DAG. It is the effect of hemispheric laterality on neurosurgical intervention that remains after holding all mediating variables constant.

The **indirect effect** of X on Y refers to the influence of X on Y that is mediated by one or more intermediate variables. It is the effect of hemispheric laterality on neurosurgical intervention that operates through other variables in the DAG.

We were interested in the direct effect of hemispheric laterality on neurosurgical intervention. We hypothesized that this relationship could be mediated by other variables that are commonly used to determine if a patient will undergo neurosurgical intervention. To estimate this effect, we included to the following covariates in our statistical model for the primary outcome: age, admission GCS, ICH location, ICH volume, and presence of IVH.

Since we felt that there were not materially different causal structures at play with our secondary outcomes, we used a similar DAG for the following outcomes: receipt of external ventricular drain (EVD), receipt of tracheostomy, days of mechanical ventilation, receipt of do-not-resuscitate (DNR) order at any point in the hospitalization, withdrawal of life-sustaining therapy (WLST) at any point in the hospitalization, and early WLST (defined as < 72 from symptom onset). We included the same covariates in our statistical models for all secondary outcomes specified above.

### Functional Outcomes

We specified the following DAG when examining functional outcomes (e.g., modified Rankin Scale and EuroQOL measures):

```{r functional outcome dag}
#| echo: false
#| warning: false
#| fig-height: 8.5
with_dir(here(), tar_load(dag_outcomes))

dag_outcomes |>
  tidy_dagitty() |>
  node_status() |>
  ggdag() +
  geom_dag_edges() +
  geom_dag_point(aes(color = status)) +
  geom_dag_label_repel(
    aes(label = label, fill = status),
    seed = 1234,
    color = "white",
    fontface = "bold"
  ) +
  scale_color_manual(values = status_colors, na.value = "grey20") +
  scale_fill_manual(values = status_colors, na.value = "grey20") +
  guides(color = "none", fill = "none") +
  theme_dag()
```

In contrast to the primary outcome analysis, where the target estimand was the direct effect of laterality on treatment decisions, the target estimand for functional outcomes was the **total effect** of hemispheric laterality. This total effect encompasses all causal pathways from laterality to functional outcome, including those mediated through neurosurgical intervention, withdrawal of life-sustaining therapy, and participation in rehabilitation. This estimand was selected because it captures the overall prognostic significance of hemispheric laterality as encountered in clinical practice — reflecting both the intrinsic biological consequences of the affected hemisphere and any downstream differences in management.

Conditioning on neurosurgical intervention — a mediator on the causal pathway from laterality to functional outcome — would block part of this clinically relevant pathway and could introduce collider stratification bias by opening non-causal associations between laterality and outcome within strata of the mediator. Furthermore, valid estimation of the controlled direct effect would require adjustment for all confounders of the mediator–outcome relationship, some of which (e.g., hematoma expansion, clinical deterioration) may themselves be influenced by the exposure, creating intermediate confounding that cannot be resolved through standard regression adjustment. Accordingly, our functional outcome models were adjusted for age, admission GCS, ICH location, ICH volume, presence of IVH, and time from symptom onset to emergency department arrival, but did not include neurosurgical intervention or other post-baseline treatment variables as covariates.

## Priors

### Prior Distribution Definitions

```{r prior definitions}
#| echo: false
#| output: false
neutral_mean <- 0
neutral_sd <- 0.35
neutral_lower_OR <- exp(neutral_mean - (1.96 * neutral_sd))
neutral_upper_OR <- exp(neutral_mean + (1.96 * neutral_sd))

left_mean <- -0.22
left_sd <- 0.175
left_lower_OR <- exp(left_mean - (1.96 * left_sd))
left_upper_OR <- exp(left_mean + (1.96 * left_sd))

right_mean <- 0.18
right_sd <- 0.175
right_lower_OR <- exp(right_mean - (1.96 * right_sd))
right_upper_OR <- exp(right_mean + (1.96 * right_sd))

flat_mean <- 0
flat_sd <- 5
flat_lower_OR <- exp(flat_mean - (1.96 * flat_sd))
flat_upper_OR <- exp(flat_mean + (1.96 * flat_sd))
```

In defining prior distributions, we generally followed suggestions from previous literature. [@zampieriUsingBayesianMethods2021; @kruschkeBayesianAnalysisReporting2021; @gelmanBayesianWorkflow2020] Specifically, we defined 4 sets of priors for each outcome:

**Neutral prior**: This prior assumed that hemispheric lateralization did not significantly affect the primary outcome and represents the **null hypothesis**. It is centered around a mean of **`r neutral_mean`** and standard deviation of **`r neutral_sd`**, corresponding to OR of **`r exp(neutral_mean)`** and 95% of the probability mass for the OR between **`r round(neutral_lower_OR, digits = 2)`** to **`r round(neutral_upper_OR, digits = 2)`**.

**Left hemisphere prior**: This prior assumes that the primary and secondary outcomes are more likely with left hemisphere injury, centering around a mean of **`r left_mean`** and standard deviation of **`r left_sd`** which corresponds to an OR of **`r round(exp(left_mean), digits = 2)`** and 95% of the probability mass for the OR between **`r round(left_lower_OR, digits = 2)`** to **`r round(left_upper_OR, digits = 2)`**.

**Right hemisphere prior**: This prior assumes that the primary and secondary outcomes are more likely with right hemisphere injury, centering around a mean of **`r right_mean`** and standard deviation of **`r right_sd`** which corresponds to an OR of **`r round(exp(right_mean), digits =2)`** and 95% of the probability mass for the OR between **`r round(right_lower_OR, digits = 2)`** to **`r round(right_upper_OR, digits =2)`**.

**Flat prior**: This prior most mimics a frequentist analysis in that the prior distribution is so wide that it essentially contributes no information to the posterior model. It is centered around a mean of **`r flat_mean`** and standard deviation of standard deviation of **`r flat_sd`** which corresponds to an OR of **`r round(exp(flat_mean), digits =2)`** and 95% of the probability mass for the OR between **`r round(flat_lower_OR, digits = 2)`** to **`r round(flat_upper_OR, digits = 20)`**.

### Prior Predictive Checks

Prior predictive checks were performed to understand the implications of the prior distributions before analyzing the data and ensuring the prior distributions represented plausible information regarding each outcome.

#### Neurosurgical Intervention

```{r prior predictive check - neurosurgery}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for neurosurgery"
#| label: "tbl-neurosurgery-prior-check"
with_dir(here(), tar_read(prior_check_neurosurgery))
```

#### Ventriculostomy

```{r prior predictive check - evd}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for ventriculostomy"
#| label: "tbl-evd-prior-check"
with_dir(here(), tar_read(prior_check_evd))
```

#### Tracheostomy

```{r prior predictive check - tracheostomy}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for tracheostomy"
#| label: "tbl-trach-prior-check"
with_dir(here(), tar_read(prior_check_tracheostomy))
```


#### Days of Mechanical Ventilation

```{r prior predictive check - vent}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for days of mechanical ventilation"
#| label: "tbl-vent-prior-check"
with_dir(here(), tar_read(prior_check_days_mechanical_ventilation))
```


#### Comfort Care

```{r prior predictive check - comfort}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for comfort care"
#| label: "tbl-comfort-prior-check"
with_dir(here(), tar_read(prior_check_comfort))
```


#### Early WLST

```{r prior predictive check - early wlst}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for early WLST"
#| label: "tbl-earlywlst-prior-check"
with_dir(here(), tar_read(prior_check_early_wlst))
```

#### DNR

```{r prior predictive check - dnr}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for DNR"
#| label: "tbl-DNR-prior-check"
with_dir(here(), tar_read(prior_check_dnr_binary))
```

#### mRS 90

```{r prior predictive check - mrs90}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for modified Rankin score at 90 days"
#| label: "tbl-mrs90-prior-check"
with_dir(here(), tar_read(prior_check_mrs_90))
```

#### EuroOOL Mobility 90

```{r prior predictive check - euro mobility}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for EuroQOL - Mobility at 90 days"
#| label: "tbl-euro-mobility-prior-check"
with_dir(here(), tar_read(prior_check_euro_mobility_90))
```

#### EuroOOL Self-Care 90

```{r prior predictive check - euro selfcare}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for EuroQOL - Self-Care at 90 days"
#| label: "tbl-euro-selfcare-prior-check"
#|
with_dir(here(), tar_read(prior_check_euro_selfcare_90))
```

#### EuroOOL Usual Activities 90

```{r prior predictive check - euro usual}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for EuroQOL - Usual Activities at 90 days"
#| label: "tbl-euro-usual-prior-check"
#|
with_dir(here(), tar_read(prior_check_euro_usual_90))
```

#### EuroOOL Pain/Discomfort 90

```{r prior predictive check - euro pain}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for EuroQOL - Pain/Discomfort at 90 days"
#| label: "tbl-euro-pain-prior-check"
#|
with_dir(here(), tar_read(prior_check_euro_pain_90))
```

#### EuroOOL Anxiety/Depression 90

```{r prior predictive check - euro anxiety}
#| echo: false
#| warning: false
#| #| tbl-cap: "Prior predictive check results for EuroQOL - Anxiety/Depression at 90 days"
#| label: "tbl-euro-anxiety-prior-check"
with_dir(here(), tar_read(prior_check_euro_anxiety_90))
```

#### EuroOOL VAS 90

```{r prior predictive check - vas}
#| echo: false
#| warning: false
#| tbl-cap: "Prior predictive check results for EuroQOL - VAS at 90 days"
#| label: "tbl-euro-vas-prior-check"
#|
with_dir(here(), tar_read(prior_check_euro_anxiety_90))
```

## Missing Data Assessment

As described in the primary manuscript, the primary analysis used multiple imputation by chained equations (MICE) under a Missing at Random (MAR) assumption. This section provides detailed reporting of missingness patterns across the analytic cohort, assessments supporting the MAR assumption, and diagnostics evaluating the quality of the imputed datasets.

### Missingness for 90-Day Outcomes by Laterality Group

Baseline covariates used in model adjustment (age, GCS, ICH volume, ICH location, IVH) had minimal missingness (<1%). However, 90-day follow-up outcomes had more substantial missingness due to loss to follow-up and differences in outcome ascertainment between the ATACH-2 and ERICH study protocols. The table below summarizes missingness for each 90-day outcome by hemispheric laterality group.

```{r missingness-summary-table}
#| echo: false
#| output: true
#| tbl-cap: "Missingness for 90-Day Outcomes by Hemispheric Laterality"
#| label: "tbl-missingness-summary"
outcome_vars <- c(
  "mrs_90", "euro_vas_90", "euro_mobility_90", "euro_selfcare_90",
  "euro_usual_90", "euro_pain_90", "euro_anxiety_90"
)

outcome_labels <- c(
  "mrs_90" = "mRS at 90 Days",
  "euro_vas_90" = "EQ-VAS at 90 Days",
  "euro_mobility_90" = "EQ-5D Mobility",
  "euro_selfcare_90" = "EQ-5D Self-Care",
  "euro_usual_90" = "EQ-5D Usual Activities",
  "euro_pain_90" = "EQ-5D Pain/Discomfort",
  "euro_anxiety_90" = "EQ-5D Anxiety/Depression"
)

missingness_long <- ich_aggressive |>
  tidyr::pivot_longer(
    cols = all_of(outcome_vars),
    names_to = "outcome",
    values_to = "value"
  ) |>
  group_by(ich_laterality, outcome) |>
  summarise(
    n = n(),
    missing = sum(is.na(value)),
    available = sum(!is.na(value)),
    .groups = "drop"
  ) |>
  mutate(
    pct_missing = scales::percent(missing / n, accuracy = 0.1),
    display = paste0(missing, "/", n, " (", pct_missing, ")"),
    outcome = outcome_labels[outcome]
  )

missingness_long |>
  select(outcome, ich_laterality, display) |>
  tidyr::pivot_wider(
    names_from = ich_laterality,
    values_from = display
  ) |>
  gt::gt() |>
  gt::cols_label(
    outcome = "Outcome",
    Left = "Left Hemisphere",
    Right = "Right Hemisphere"
  ) |>
  gt::tab_source_note(
    source_note = "Values are n missing / N total (%). mRS indicates modified Rankin Scale; EQ-VAS, European Quality of Life Visual Analog Scale; EQ-5D, European Quality of Life-5 Dimensions."
  )
```

### Missingness Patterns by Laterality and Study

The following figures show the percentage of missing data across all analysis variables, stratified by hemispheric laterality and by study. Missingness patterns were largely driven by differences between the ATACH-2 and ERICH study protocols rather than by hemispheric laterality.

```{r missing-data-by-laterality}
#| echo: false
#| output: true
#| fig-cap: "Percentage of missing data by hemispheric laterality"
#| label: "fig-missing-by-laterality"
#| fig-height: 8
with_dir(here(), tar_read(missing_data_by_ich_laterality))
```

```{r missing-data-by-study}
#| echo: false
#| output: true
#| fig-cap: "Percentage of missing data by study"
#| label: "fig-missing-by-study"
#| fig-height: 8
with_dir(here(), tar_read(missing_data_by_study))
```

### Assessment of Missing at Random Assumption

To assess whether missingness in 90-day outcomes was associated with key baseline covariates, we generated shadow plots comparing the distribution of each covariate among patients with observed versus missing values. If the distributions are similar, this supports the MAR assumption — that missingness is conditionally independent of the unobserved outcome given the observed covariates.

#### mRS at 90 Days

```{r shadow-age-mrs}
#| echo: false
#| output: true
#| fig-height: 6
with_dir(here(), tar_read(missingness_check_age_mrs_90))
```

```{r shadow-gcs-mrs}
#| echo: false
#| output: true
#| fig-height: 6
with_dir(here(), tar_read(missingness_check_gcs_baseline_mrs_90))
```

```{r shadow-volume-mrs}
#| echo: false
#| output: true
#| fig-height: 6
with_dir(here(), tar_read(missingness_check_ich_volume_baseline_mrs_90))
```

```{r shadow-time-mrs}
#| echo: false
#| output: true
#| fig-height: 6
with_dir(here(), tar_read(missingness_check_time_symptoms_to_ed_mrs_90))
```

#### EQ-VAS at 90 Days

```{r shadow-age-vas}
#| echo: false
#| output: true
#| fig-height: 6
with_dir(here(), tar_read(missingness_check_age_euro_vas_90))
```

```{r shadow-gcs-vas}
#| echo: false
#| output: true
#| fig-height: 6
with_dir(here(), tar_read(missingness_check_gcs_baseline_euro_vas_90))
```

```{r shadow-volume-vas}
#| echo: false
#| output: true
#| fig-height: 6
with_dir(here(), tar_read(missingness_check_ich_volume_baseline_euro_vas_90))
```

```{r shadow-time-vas}
#| echo: false
#| output: true
#| fig-height: 6
with_dir(here(), tar_read(missingness_check_time_symptoms_to_ed_euro_vas_90))
```

### Imputation Diagnostics

The following figure presents diagnostic checks for the MICE imputation procedure, comparing the distributions of observed data against imputed values across the 20 imputed datasets. Consistency between the observed and imputed distributions supports the plausibility of the imputation model.

```{r imputation-diagnostics}
#| echo: false
#| output: true
#| fig-cap: "Imputation diagnostic plots comparing observed and imputed distributions"
#| label: "fig-imputation-diagnostics"
#| fig-height: 12
with_dir(here(), tar_read(imputed_visualizations))
```

## Posterior Simulation Diagnostics

Posterior Simulation Checks
To ensure the reliability and validity of our Bayesian models, we performed several posterior simulation checks.

Chain Convergence: We assessed the convergence of the Markov Chain Monte Carlo (MCMC) chains by visually inspecting trace plots for each parameter. The trace plots showed that the chains were well-mixed and stationary, indicating that they had converged to a stable posterior distribution. Additionally, we calculated the R-hat statistic for all parameters, ensuring that all values were below 1.01, which confirms the convergence of the chains.

Posterior Predictive Checks: We conducted posterior predictive checks to evaluate the model's fit to the data. This involved simulating data from the posterior predictive distribution and comparing it to the observed data. The close correspondence between the simulated and observed data confirmed that our models provided a good fit to the underlying data structure.

### MCMC Chain Convergence and Resolution

#### Neurosurgical Intervention

```{r posterior diagnostics neurosurgery}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for Neurosurgical Intervention"
#| label: "tbl-neurosurgical-posterior-diag"
#|
with_dir(here(), tar_read(diag_posterior_neutral_neurosurgery))
```

#### Ventriculostomy

```{r posterior diagnostics evd}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for Ventriculostomy"
#| label: "tbl-evd-posterior-diag"
with_dir(here(), tar_read(diag_posterior_neutral_evd))
```

#### Tracheostomy

```{r posterior diagnostics trach}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for Tracheostomy"
#| label: "tbl-trach-posterior-diag"
with_dir(here(), tar_read(diag_posterior_neutral_tracheostomy))
```

#### Days of Mechanical Ventilation

```{r posterior diagnostics vent}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for Days of Mechanical ventilation"
#| label: "tbl-vent-posterior-diag"
#|
with_dir(here(), tar_read(diag_posterior_neutral_days_mechanical_ventilation))
```

#### Comfort Care

```{r posterior diagnostics comfort}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for comfort care"
#| label: "tbl-comfort-posterior-diag"
#|
with_dir(here(), tar_read(diag_posterior_neutral_comfort))
```

#### Early WLST

```{r posterior diagnostics early wlst}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for early WLST"
#| label: "tbl-earlywlst-posterior-diag"
#|
with_dir(here(), tar_read(diag_posterior_neutral_early_wlst))
```

#### DNR

```{r posterior diagnostics dnr}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for DNR"
#| label: "tbl-dnr-posterior-diag"
#|
with_dir(here(), tar_read(diag_posterior_neutral_dnr_binary))
```

#### mRS 90

```{r posterior diagnostics mrs}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for modified Rankin score at 90 days"
#| label: "tbl-mrs90-posterior-diag"
#|
with_dir(here(), tar_read(diag_posterior_mrs_90_minimal))
```

#### EuroQOL Mobility 90

```{r posterior diagnostics euro mobility}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for EuroQOL - Mobility at 90 days"
#| label: "tbl-euro-mobility-posterior-diag"
with_dir(here(), tar_read(diag_posterior_neutral_euro_mobility_90_canonical))
```

#### EuroQOL Self-Care 90

```{r posterior diagnostics euro selfcare}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for EuroQOL - Self-Care at 90 days"
#| label: "tbl-euro-selfcaremobility-posterior-diag"
with_dir(here(), tar_read(diag_posterior_neutral_euro_selfcare_90_canonical))
```

#### EuroQOL Usual Activities 90

```{r posterior diagnostics euro usual}
#| echo: false
#| fig-height: 10
#| #| tbl-cap: "Posterior diagnostic checks for EuroQOL - Usual Activities at 90 days"
#| label: "tbl-euro-usual-posterior-diag"
with_dir(here(), tar_read(diag_posterior_neutral_euro_usual_90_canonical))
```

#### EuroQOL Pain/Discomfort 90

```{r posterior diagnostics euro pain}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for EuroQOL - Pain/Discomfort at 90 days"
#| label: "tbl-euro-pain-posterior-diag"
with_dir(here(), tar_read(diag_posterior_neutral_euro_pain_90_canonical))
```

#### EuroQOL Anxiety/Depression 90

```{r posterior diagnostics euro anxiety}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for EuroQOL - Anxiety/Depression at 90 days"
#| label: "tbl-euro-anxiety-posterior-diag"
with_dir(here(), tar_read(diag_posterior_neutral_euro_anxiety_90_canonical))
```

#### EuroQOL VAS 90

```{r posterior diagnostics euro vas}
#| echo: false
#| fig-height: 10
#| tbl-cap: "Posterior diagnostic checks for EuroQOL - VAS at 90 days"
#| label: "tbl-euro-vas-posterior-diag"
with_dir(here(), tar_read(diag_posterior_vas_90))
```

### Posterior predictive checks

#### Neurosurgical Intervention

```{r posterior predictions neurosurgery}
#| echo: false
#| tbl-cap: "Posterior predictive checks for neurosurgical intervention"
#| label: "tbl-neurosurgery-posterior-checks"
with_dir(here(), tar_read(post_pred_check_neurosurgery))
```

#### Ventriculostomy

```{r posterior predictions evd}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for ventriculostomy"
#| label: "tbl-evd-posterior-checks"
with_dir(here(), tar_read(post_pred_check_evd))
```

#### Tracheostomy

```{r posterior predictions trach}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for tracheostomy"
#| label: "tbl-trach-posterior-checks"
with_dir(here(), tar_read(post_pred_check_tracheostomy))
```

#### Days of Mechanical Ventilation

```{r posterior predictions vent}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for days of mechanical ventilation"
#| label: "tbl-vent-posterior-checks"
with_dir(here(), tar_read(post_pred_check_days_mechanical_ventilation))
```

#### Comfort Care

```{r posterior predictions comfort}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for comfort care"
#| label: "tbl-comfort-posterior-checks"
with_dir(here(), tar_read(post_pred_check_comfort))
```

#### Early WLST

```{r posterior predictions early wlst}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for early WLST"
#| label: "tbl-earlywlst-posterior-checks"
with_dir(here(), tar_read(post_pred_check_early_wlst))
```

#### DNR

```{r posterior predictions dnr}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for DNR"
#| label: "tbl-dnr-posterior-checks"
with_dir(here(), tar_read(post_pred_check_dnr_binary))
```

#### mRS 90

```{r posterior predictions mrs}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for modified Rankin Score 90 days"
#| label: "tbl-mrs90-posterior-checks"
with_dir(here(), tar_read(post_pred_check_neutral_mrs_90_canonical))
```

#### EuroQOL Moblity 90

```{r posterior predictions euro mobility}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for EuroQOL - Mobility 90 days"
#| label: "tbl-euro-mobility-checks"
with_dir(here(), tar_read(post_pred_check_neutral_euro_mobility_90_canonical))
```

#### EuroQOL Self-Care 90

```{r posterior predictions euro selfcare}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for EuroQOL - Self-Care 90 days"
#| label: "tbl-euro-selfcare-checks"
with_dir(here(), tar_read(post_pred_check_neutral_euro_selfcare_90_canonical))
```

#### EuroQOL Usual Activities 90

```{r posterior predictions euro usual}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for EuroQOL - Usual Activities 90 days"
#| label: "tbl-euro-usual-checks"
with_dir(here(), tar_read(post_pred_check_neutral_euro_usual_90_canonical))
```

#### EuroQOL Pain/Discomfort 90

```{r posterior predictions euro pain}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for EuroQOL - Pain/Discomfort 90 days"
#| label: "tbl-euro-pain-checks"
with_dir(here(), tar_read(post_pred_check_neutral_euro_pain_90_canonical))
```

#### EuroQOL Anxiety/Despression 90

```{r posterior predictions euro anxiety}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for EuroQOL - Anxiety/Depression 90 days"
#| label: "tbl-euro-anxiety-checks"
with_dir(here(), tar_read(post_pred_check_neutral_euro_anxiety_90_canonical))
```

#### EuroQOL VAS 90

```{r posterior predictions euro vas}
#| echo: false
#| tbl-cap: "Posterior diagnostic checks for EuroQOL - VAS 90 days"
#| label: "tbl-euro-vas-checks"
with_dir(here(), tar_read(post_pred_check_vas_90))
```

## Sensitivity Analyses

To assess the robustness of our findings for the primary outcome, we conducted several sensitivity analyses. First, we repeated the primary analysis using three alternative prior distributions: a left-hemisphere prior, a right-hemisphere prior, and a flat (minimally informative) prior. This allowed us to evaluate the influence of different initial beliefs on the resulting posterior distributions. Second, we explored an alternative covariate adjustment set based on a minimal adjustment set identified from our directed acyclic graphs (DAGs) to ensure our results were not dependent on the specific model specification. Third, as described in the primary manuscript, the primary analysis used MICE-imputed data; we performed a complete-case sensitivity analysis (excluding patients with missing outcomes) to confirm that our findings were not unduly influenced by the imputation assumptions. Finally, to evaluate the potential impact of center-level clustering, we performed a site-level random-effect sensitivity analysis restricted to the ATACH-2 cohort, where hospital site identifiers were available. The results of these sensitivity analyses were consistent with our primary findings, demonstrating the robustness of our conclusions.

### Exploratory Subgroup Interaction Analysis

To evaluate whether the association between hemispheric laterality and neurosurgical intervention varied across clinically relevant subgroups, we fitted Bayesian logistic regression models that included formal interaction terms between ICH laterality and each subgroup variable. Two interaction models were fitted: (1) laterality × ICH location (basal ganglia, lobar, thalamic) and (2) laterality × study (ATACH-2, ERICH). Both models used the neutral prior under the adjusted covariate set and were fitted to complete-case data.

For each subgroup, we report: the subgroup-specific adjusted odds ratio (aOR) for the laterality association with 95% CrI, posterior probabilities of the aOR exceeding 1 and 1.2, the proportion of the posterior within the ROPE, and — critically — the formal interaction effect expressed as a Ratio of Odds Ratios (ROR). The ROR quantifies how many times larger (or smaller) the laterality effect is in a given subgroup compared to the reference subgroup. An ROR of 1 indicates no difference in the laterality association across subgroups, while values substantially above or below 1 suggest effect modification.

This approach avoids the well-known pitfalls of comparing isolated subgroup point estimates, which can yield misleading conclusions about heterogeneity. Instead, the interaction ROR directly tests whether the laterality effect differs between subgroups and provides a formal probabilistic assessment of the degree of heterogeneity.

```{r subgroup-interaction-analysis}
#| echo: false
#| output: true
#| tbl-cap: "Subgroup Analysis of Hemispheric Laterality and Neurosurgical Intervention with Formal Interaction Tests"
#| label: "tbl-subgroup-interactions"

with_dir(here(), tar_read(table_subgroups))
```

The laterality association with neurosurgical intervention was directionally consistent across ICH location and study subgroups. While the point estimates suggested a potentially larger laterality effect for lobar hemorrhages relative to basal ganglia hemorrhages, the interaction ROR credible intervals were wide and included 1, indicating that the observed subgroup differences were not statistically distinguishable. Similarly, the laterality effect was consistent between the ATACH-2 and ERICH cohorts, with no evidence of meaningful effect modification by study source.

### Table 2 with Different Priors

#### Left Hemisphere Prior

```{r}
#| echo: false
#| #| tbl-cap: "Primary and Secondary Outcomes of Interest with Left Hemisphere Prior"
#| label: "tbl-2-left-prior"
with_dir(here(), tar_read(table2_left))
```

#### Right Hemisphere Prior

```{r}
#| echo: false
#| #| tbl-cap: "Primary and Secondary Outcomes of Interest with Right Hemisphere Prior"
#| label: "tbl-2-right-prior"
#
with_dir(here(), tar_read(table2_right))
```

#### Flat Prior

```{r}
#| echo: false
#| #| tbl-cap: "Primary and Secondary Outcomes of Interest with Flat Prior"
#| label: "tbl-2-flat-prior"
with_dir(here(), tar_read(table2_flat))
```

### Table 4 with Different Priors

#### Left Hemisphere Prior

```{r}
#| echo: false
#| #| tbl-cap: "Functional  Outcomes of Interest with Left Hemisphere Prior"
#| label: "tbl-4-left-prior"
#|
with_dir(here(), tar_read(table4_left))
```

#### Right Hemisphere Prior

```{r}
#| echo: false
#| #| tbl-cap: "Functional  Outcomes of Interest with Right Hemisphere Prior"
#| label: "tbl-4-right-prior"
#|
with_dir(here(), tar_read(table4_right))
```

#### Flat Prior

```{r}
#| echo: false
#| #| tbl-cap: "Functional  Outcomes of Interest with Flat Prior"
#| label: "tbl-4-flat-prior"
#|
with_dir(here(), tar_read(table4_flat))
```

### Complete-Case Sensitivity Analysis

The primary analysis used multiple imputation by chained equations (MICE) with 20 imputed datasets, as described in the main manuscript. As a sensitivity analysis, we repeated the primary models using a complete-case approach (excluding patients with missing outcomes) under the neutral prior with the adjusted covariate set. Three models were fitted to the complete-case data using the following outcomes: neurosurgical intervention, external ventricular drain, and modified Rankin score at 90 days. All models were fitted using Markov Chain Monte Carlo (MCMC) with 4 chains and 5,000 iterations per chain (including a warmup period of 1,000 iterations).

```{r imputed}
#| echo: false
#| tbl-cap: "Outcomes of Interest Using Complete-Case Analysis (Sensitivity)"
#| label: "tbl-complete-case-sensitivity"
#|
with_dir(here(), tar_read(table_imp))
```

### Site-Level Clustering Sensitivity Analysis

Neurosurgical decision-making is highly influenced by local practice patterns, institutional resources, and surgeon availability. Our primary models included a random intercept for study (ATACH-2 versus ERICH) to account for systematic differences between the two cohorts, but this two-level random effect is unlikely to fully capture the heterogeneity in clinical practice across the many individual hospital sites within each study.

Ideally, we would have included a random intercept for hospital site in all models. However, site-level identifiers were only available for the ATACH-2 cohort and were not included in the ERICH dataset. To assess the potential impact of site-level clustering on our primary finding, we performed a sensitivity analysis restricted to the ATACH-2 cohort, where we fitted two versions of the primary model for neurosurgical intervention:

1. **Base model**: The same model specification as the primary analysis (adjusted for ICH location, age, GCS, ICH volume, IVH, and an ICH laterality-by-location interaction) but without any random intercept, since ATACH-2 is a single study.

2. **Site random-effect model**: The same model specification with the addition of a random intercept for hospital site, allowing the baseline probability of neurosurgical intervention to vary across sites.

Both models were fitted using multiple imputation (20 imputed datasets) with the neutral prior under the adjusted covariate set. The table below compares the posterior estimates for the association between hemispheric laterality and neurosurgical intervention from both models.

```{r atach-site-sensitivity}
#| echo: false
#| output: true
#| tbl-cap: "ATACH-2 Sensitivity Analysis: Association of Hemispheric Laterality with Neurosurgical Intervention With and Without Site-Level Random Intercept"
#| label: "tbl-atach-site-sensitivity"
with_dir(here(), tar_read(table_2_atach))
```

Inclusion of the site-level random intercept slightly attenuated the point estimate for the laterality association and widened the credible interval, which is expected given the additional variance component and the smaller sample size of the ATACH-2 cohort alone. However, the direction, magnitude, and posterior probability of the association remained consistent with the primary analysis, suggesting that site-level clustering does not materially alter the principal finding.